## Load and prepare BCR data

```{r}
# samples to load
subjects <- c(paste0("C", 1:6), paste0("P", 2:8))
bcr_data_list <- list()

# Loop through the sample names, load the airrflow-processed data, and create unique id
for (subject in subjects) {
  # Read the data
  bcr_data <- airr::read_rearrangement(file.path(path_airr_BCR, "all_reps_clone_report", "repertoires", paste0(subject, "__clone-pass.tsv")))
  
  # Simplify sample_id
  bcr_data$sample_id <- gsub("_BCR", "", bcr_data$sample_id)
  
  # Assign unique cell_id
  bcr_data$cell_id_unique <- paste0(bcr_data$sample_id, "_", bcr_data$cell_id)
  bcr_data$cell_id_unique <- gsub("-.*", "", bcr_data$cell_id_unique)
  
  # Assign unique clone_id
  bcr_data$clone_id_unique <- paste0(bcr_data$subject_id, "_", bcr_data$clone_id)
  
  # Dynamically assign the transformed data to a variable
  bcr_data_list[[subject]] <- bcr_data
}

# Combine all data frames in the list into a single data frame
bcr_data <- do.call(rbind, bcr_data_list)

# Combine split aliquots for same sample
bcr_data <- bcr_data %>% mutate(sample_final = case_when(
                                            sample_id %in% c("C4_1_1_BLD", "C4_2_1_BLD") ~ "C4_1-2_1_BLD",
                                            sample_id %in% c("C5_1_2_BLD", "C5_2_2_BLD") ~ "C5_1-2_2_BLD",
                                            sample_id %in% c("C6_1_1_BLD", "C6_2_1_BLD") ~ "C6_1-2_1_BLD", 
                                            sample_id %in% c("P3_1_5_BLD", "P3_2_5_BLD") ~ "P3_1-2_5_BLD", 
                                            sample_id %in% c("P4_1_8_BLD", "P4_2_8_BLD") ~ "P4_1-2_8_BLD", 
                                            sample_id %in% c("P5_1_7_BLD", "P5_2_7_BLD") ~ "P5_1-2_7_BLD", 
                                            sample_id %in% c("P6_1_1_BLD", "P6_2_1_BLD") ~ "P6_1-2_1_BLD", 
                                            sample_id %in% c("P7_1_3_BLD", "P7_2_3_BLD") ~ "P7_1-2_3_BLD",
                                            TRUE ~ sample_id))
bcr_data$sample_id <- bcr_data$sample_final
bcr_data$sample_final <- NULL

# remove columns clone_size_count and clone_size_freq from airrflow
# these will be recalculated after matching BCR with GEX
bcr_data$clone_size_count <- NULL
bcr_data$clone_size_freq <- NULL

# transform mu_freq to numeric
bcr_data$mu_freq <- as.numeric(bcr_data$mu_freq)

# remove objects
rm(subject, subjects, bcr_data_list)
```

## Add and prepare metadata

```{r}
# add column showing controls
bcr_data <- bcr_data %>% mutate(HIV_status = case_when(
    subject_id %in% c("C1", "C2", "C3", "C4", "C5", "C6") ~ "control",
    subject_id %in% c("P1", "P2", "P3", "P4", "P5", "P6", "P7", "P8") ~ "PWH",
    TRUE ~ ""
  ))

# group isotypes
bcr_data <- bcr_data %>% mutate(isotype_group = case_when(
  isotype %in% c("IGHG1", "IGHG2", "IGHG3", "IGHG4") ~ "IgG", 
  isotype %in% c("IGHA1", "IGHA2") ~ "IgA", 
  isotype == "IGHM" ~ "IgM",
  isotype == "IGHD" ~ "IgD", 
  isotype == "IGHE" ~ "IgE",
  TRUE ~ isotype
))

# reorder factor variables
bcr_data$tissue <- factor(bcr_data$tissue, levels = c("PBMC", "CSF"))
bcr_data$isotype_group <- factor(bcr_data$isotype_group, levels = c("IgD", "IgM", "IgG", "IgA", "IgE"))
bcr_data$isotype <- factor(bcr_data$isotype, levels = c("IGHD", "IGHM", "IGHG1", "IGHG2", "IGHG3", "IGHG4", "IGHA1", "IGHA2", "IGHE"))

# calculate CDR3 length
bcr_data <- bcr_data %>% mutate(cdr3_length_aa = nchar(cdr3)/3)

# save table for BCR matching in GEX
write.csv(bcr_data, row.names = FALSE, file.path(path_objs, "bcr_data_gex.csv"))
```

## Add GEX information

This part of the analysis is done after B cell annotation in GEX analysis.

```{r}
# add B cell subclustering annotations
cell.annotation.B <- as.character(Idents(gex_B_ART))

# match barcodes to find the position of GEX cells in BCR data
match.index <- match(bcr_data$cell_id_unique, Cells(gex_B_ART))

# B cell annotations
# add annotations of B cell subclustering
bcr_data <- bcr_data %>% mutate(gex_annotation = ifelse(!is.na(match.index), cell.annotation.B[match.index], NA))

# remove unnecessary objects
rm(cell.annotation.B, match.index)
```

### GEX-BCR overlap

```{r}
# explore GEX-BCR overlap
ggVennDiagram(
  x = list("BCR" = unique(bcr_data$cell_id_unique),
           "GEX_B" = Cells(gex_B_ART), 
           "GEX_All" = Cells(gex_ART))) +
  scale_fill_distiller(palette = "Greys", direction = 1) +
  ggtitle("Annotated BCRs - All tissues") + NoLegend()
```

### Filter only for B cells from GEX that have BCR

```{r}
# filter intersection of BCR and GEX
bcr_data <- bcr_data %>% dplyr::filter(!is.na(gex_annotation))
```

### Refine B cell annotations

The following annotations based on GEX and isotypes are biologically meaningful 
- Naive IgM+/IgD+ cells and Naive IgM-IgD-
- Memory IgD+, Memory IgM+ (unswitched), Memory IgG+/IgA+/IgE+

```{r}
# re-annotate B cells based on isotypes
bcr_data <- bcr_data %>%
  mutate(gex_annotation_final = case_when(
  gex_annotation == "Naive" & isotype %in% c("IGHD", "IGHM") ~ "Naive", 
  gex_annotation == "Naive" & !isotype %in% c("IGHD", "IGHM") ~ "Naive IgD/IgM-", 
  gex_annotation == "Resting Memory" & isotype == "IGHD" ~ "MBC IgD+",
  gex_annotation == "Resting Memory" & isotype == "IGHM" ~ "MBC class-unswitched",
  gex_annotation == "Resting Memory" & !isotype %in% c("IGHD", "IGHM") ~ "MBC class-switched",
  TRUE ~ gex_annotation)) %>% 
  group_by(cell_id_unique) %>%
  mutate(gex_annotation_final = gex_annotation_final[locus == "IGH"])

# reorder annotations
bcr_data$gex_annotation_final <- factor(bcr_data$gex_annotation_final, 
                                  levels = c("Naive", "Naive IgD/IgM-", "Atypical Memory", "MBC IgD+", "MBC class-unswitched", "MBC class-switched", "Plasmablasts"))
```

## Call clones with B cells that were kept

```{r}
# remove clone_id information provided by airrflow
bcr_data$clone_id <- NULL
bcr_data$clone_id_unique <- NULL
```

### Find clonal threshold

```{r}
# Define the subject IDs
subject_ids <- c("P2", "P3", "P4", "P5", "P6", "P7", "P8", "C1", "C2", "C3", "C4", "C5", "C6")

# Initialize objects to store results
threshold_table <- data.frame(subject_id = character(), threshold = numeric(), stringsAsFactors = FALSE)
plot_list <- list()

# Calculate cross-subject distribution of distance to nearest
cat("Calculating cross-subject distribution of distance to nearest...\n")
dist_crossSubj <- distToNearest(dplyr::filter(bcr_data, locus == "IGH"), nproc = 4, cross = "subject_id")

# Loop through each subject_id
for (subject in subject_ids) {
  cat("Processing subject_id:", subject, "\n")
  
  # Calculate distance to nearest for the current subject_id
  dist_nearest <- distToNearest(dplyr::filter(bcr_data, locus == "IGH", subject_id == subject))

  # Find the threshold using GMM method
  threshold_output <- shazam::findThreshold(
    dist_nearest$dist_nearest,
    method = "gmm",
    model = "gamma-norm",
    cross = dist_crossSubj$cross_dist_nearest,
    cutoff = "user",
    spc = 0.995
  )

  # Extract the threshold value
  threshold_value <- threshold_output@threshold

  # Store the threshold in the table
  threshold_table <- rbind(threshold_table, data.frame(subject_id = subject, threshold = threshold_value))

  # Create and store the plot
  plot <- plot(threshold_output, binwidth = 0.02,
               cross = dist_crossSubj$cross_dist_nearest, silent = TRUE) +
    theme(axis.title = element_text(size = 18)) +
    ggtitle(paste("Threshold for", subject))
  plot_list[[subject]] <- plot
}

# Print the threshold table
cat("\nThresholds calculated for all subject_ids:\n")
print(threshold_table)

# remove objects
rm(subject, subject_ids, threshold_output, threshold_value, dist_nearest, dist_crossSubj, plot)
```

### Define clonal groups in filtered data

```{r}
# call clones using hierarchicalClones
# based on the cross-subject distribution a theshold of 0.12 is selected
bcr_data <- hierarchicalClones(bcr_data,
                              cell_id = "cell_id_unique",
                              threshold = 0.12,
                              only_heavy = TRUE, split_light = TRUE,
                              summarize_clones = FALSE,
                              fields = "subject_id")

# create unique clone_id
bcr_data$clone_id_unique <- paste0(bcr_data$subject_id, "_", bcr_data$clone_id)
```

### Calculate clone size and identify shared clones

```{r}
# add clone_size column
bcr_data <-  bcr_data %>%
                group_by(clone_id_unique) %>%
                mutate(clone_size = n_distinct(cell_id_unique)) %>%
                ungroup()

# identify expanded clones
expanded <- countClones(filter(bcr_data, locus == "IGH"), clone = "clone_id_unique") %>% dplyr::filter(seq_count > 1) %>% pull(clone_id_unique)

# identify clones in blood and CSF separately
clones_blood <- bcr_data %>% dplyr::filter(locus == "IGH" & tissue == "PBMC") %>% pull(clone_id_unique) %>% unique()
clones_CSF <- bcr_data %>% dplyr::filter(locus == "IGH" & tissue == "CSF") %>% pull(clone_id_unique) %>% unique()

# identify shared clones across tissues and create new meta column
shared_clones <- intersect(clones_blood, clones_CSF)
bcr_data <- bcr_data %>% mutate(shared_BLD_CSF = clone_id_unique %in% shared_clones)

# add column showing expanded cells
bcr_data <- bcr_data %>% mutate(expanded = (shared_BLD_CSF == TRUE) | (shared_BLD_CSF == FALSE & clone_id_unique %in% expanded))

# remove unnecessary objects
rm(expanded, shared_clones, clones_blood, clones_CSF)
```

### Save final object

```{r}
write.csv(bcr_data, row.names = FALSE, file.path(path_objs, "bcr_data_filtered.csv"))
```

### Subset data for heavy chains

```{r}
bcr_heavy <- bcr_data %>% dplyr::filter(locus == "IGH")
```

## BCR repertoire characteristics

### Number of B cells per group

```{r}
# calculate nb of sequences per subject_id and tissue
bcr_heavy %>% group_by(subject_id, tissue, gex_annotation_final) %>% summarize(n_seq = n())
```

### CDR3 Length

#### CDR3 distribution

```{r}
# calculate cell frequency per CDR3 length
cdr3_length_freq <- bcr_heavy %>%
  group_by(cdr3_length_aa, tissue, HIV_status) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(tissue, HIV_status) %>%
  mutate(freq = count / sum(count)) %>%
  as.data.frame() %>%
  mutate(cdr3_length_aa = as.factor(cdr3_length_aa))
```

```{r}
# Compare CDR3 length across all clusters
bcr_heavy %>%
  group_by(subject_id, tissue, HIV_status) %>% 
  summarize(mean_cdr3_length = mean(cdr3_length_aa), .groups = "drop") %>%
  group_by(tissue, HIV_status) %>%
  summarize(median_of_average_CDR3 = median(mean_cdr3_length),
            Q1 = quantile(mean_cdr3_length, probs = 0.25),
            Q3 = quantile(mean_cdr3_length, probs = 0.75),
            .groups = "drop")

# Compare CDR3 length between clusters
bcr_heavy %>%
  dplyr::filter(gex_annotation_final %in% c("MBC class-unswitched", "MBC class-switched")) %>%
  group_by(subject_id, tissue, HIV_status, gex_annotation_final) %>% 
  summarize(mean_cdr3_length = mean(cdr3_length_aa))
```

### Light chain usage

```{r}
# Light chain usage and capture per participant
light_usage_all <- bcr_data %>% 
  group_by(cell_id_unique, locus, subject_id, tissue, HIV_status) %>% 
  summarize(count = n()) %>% 
  spread(key = locus, value = count, fill = 0) %>%
  mutate(IGK = ifelse(IGK > 1, 1, IGK), 
         IGL = ifelse(IGL > 1, 1, IGL)) %>%
  mutate(light_group = case_when(
    IGK == 0 & IGL == 0 ~ "none", 
    IGK == 1 & IGL == 0 ~ "kappa",
    IGK == 0 & IGL == 1 ~ "lambda", 
    IGK == 1 & IGL == 1 ~ "both",
    TRUE ~ NA
  )) %>%
  group_by(subject_id, tissue, HIV_status, light_group) %>%
  summarize(count = n()) %>%
  group_by(subject_id, HIV_status, tissue) %>%
  mutate(fraction = count / sum(count)) %>%
  ungroup()

# light chain usage comparison across groups
ratio_data <- bcr_data %>%
  group_by(tissue, HIV_status, subject_id, locus) %>%
  summarize(count = n(), .groups = "drop") %>%
  spread(key = locus, value = count, fill = 0) %>%
  mutate(lambda_fraction = IGL / (IGK + IGL), 
         kappa_fraction = IGK / (IGK + IGL),
         kappa_lambda = IGK / IGL)
```

## Clonal distribution

```{r}
# clonal_distribution per subject and tissue in PWH
clones_sizes_PWH <- countClones(dplyr::filter(bcr_heavy, expanded == TRUE & HIV_status == "PWH"), groups = c("subject_id", "tissue"))
clones_sizes_PWH$tissue <- factor(clones_sizes_PWH$tissue, levels = c("PBMC", "CSF"))

p1 <- ggplot(clones_sizes_PWH, aes(x = seq_count)) +
  geom_bar() +
  facet_grid(tissue ~ subject_id, drop = FALSE) +
  labs(x = "Sequences per clone", title = "Clone size distribution in PWH (clone size >1)") +
  theme_bw()

# clonal_distribution per subject and tissue in controls
clones_sizes_controls <- countClones(dplyr::filter(bcr_heavy, expanded == TRUE & HIV_status == "control"), groups = c("subject_id", "tissue"))
clones_sizes_controls$tissue <- factor(clones_sizes_controls$tissue, levels = c("PBMC", "CSF"))

p2 <- ggplot(clones_sizes_controls, aes(x = seq_count)) +
  geom_bar() +
  facet_grid(tissue ~ subject_id, drop = FALSE) +
  labs(x = "Sequences per clone", title = "Clone size distribution in controls (clone size >1)") +
  theme_bw()

p1 / p2

# remove objects
rm(clones_sizes_controls, clones_sizes_PWH)
```

## Clonal expansion

### Qualitative assessment of clonal expansion

```{r}
# Summarize data
bcr_heavy %>%
  group_by(clone_id_unique, subject_id, HIV_status, tissue) %>%
  summarize(
    total_cells    = n_distinct(cell_id_unique),
    expanded_cells = sum(expanded, na.rm = TRUE),
    shared_cells   = sum(shared_BLD_CSF, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  group_by(subject_id, HIV_status, tissue) %>%
  summarize(
    n_clones     = n_distinct(clone_id_unique),
    n_singletons = sum(total_cells == 1),
    n_expanded   = sum(expanded_cells > 0), 
    n_shared     = sum(shared_cells > 0)
  ) %>%
  mutate(non_singletons = n_clones - n_singletons,
         percent_non_singleton = round(100 * non_singletons / n_clones, 1))
```

### Rank abundance curve

```{r}
# plot the rank-abundance curves per subject_id and tissue in blood
abund_subject_BLD <- estimateAbundance(dplyr::filter(bcr_heavy, tissue == "PBMC"),
                                       group = "subject_id", nboot = 200, min_n = 30)

plot(abund_subject_BLD) + facet_wrap("subject_id", ncol = 6) + 
  ggtitle(paste0("Rank-abundance curve in PBMC (N = ", min(abund_subject_BLD@n), ")")) +
  theme(plot.title = element_text(hjust = 0.5, size = 14))


# plot the rank-abundance curves per subject_id and tissue in CSF
# bootstrapping is inappropriate for CSF given the low number of sequences 
# set the min threshold to 5 to enable the use of the function - ignore error zone
abund_subject_CSF <- estimateAbundance(dplyr::filter(bcr_heavy, tissue == "CSF"),
                                       group = "subject_id", nboot = 200, min_n = 5)

plot(abund_subject_CSF) + facet_wrap("subject_id", ncol = 3) +
  ggtitle(paste0("Rank-abundance curve in CSF (N = ", min(abund_subject_CSF@n), ")")) +
  theme(plot.title = element_text(hjust = 0.5, size = 14))

# remove objects
rm(abund_subject_BLD, abund_subject_CSF)
```

### Diversity analysis

```{r}
# plot diversity curves per subject_id and tissue in blood
div_PBMC <- alphaDiversity(dplyr::filter(bcr_heavy, tissue == "PBMC"),
                      group = "subject_id", nboot = 200, min_n = 30)

plot(div_PBMC, silent=T) + facet_wrap(~ factor(subject_id, levels = c(paste0("P", 2:8), paste0("C", 1:6))), ncol = 7) +
  ggtitle(paste0("Diversity curve in PBMC (N = ", min(div_PBMC@n), ")")) +
  theme(plot.title = element_text(hjust = 0.5, size = 14))

# plot diversity curves per subject_id and tissue in CSF
# bootstrapping is inappropriate for CSF given the low number of sequences 
# set the min threshold to 5 to enable the use of the function - ignore error zone
div_CSF <- alphaDiversity(dplyr::filter(bcr_heavy, tissue == "CSF"),
                      group = "subject_id", nboot = 200, min_n = 5)

plot(div_CSF, silent=T) + facet_wrap(~ factor(subject_id, levels = c(paste0("P", 2:8), paste0("C", 1:6))), ncol = 7) +
  ggtitle(paste0("Diversity curve in CSF (N = ", min(div_CSF@n), ")")) +
  theme(plot.title = element_text(hjust = 0.5, size = 14))

# remove objects
rm(div_PBMC, div_CSF)
```

## Shared Clones

```{r}
# list specific shared clones
bcr_heavy %>%
  distinct(subject_id, tissue, cell_id_unique, clone_id_unique) %>%
  group_by(subject_id, tissue, clone_id_unique) %>%
  summarize(n = n()) %>%
  spread(tissue, n, fill = 0) %>%
  dplyr::filter(PBMC > 0 & CSF > 0) %>%
  kable() %>% kable_styling("striped")
```

## Isotype analysis

### Barplots

#### By subject

```{r}
# isotypes and subtypes
plot_pcts(pcts = calc_pcts(data = dplyr::filter(bcr_heavy, !is.na(isotype)),
                           meta_group_by = c("subject_id", "tissue", "HIV_status"),
                           focus_group = "isotype"), plot_type = "All", plot_value = "Isotypes",
          x_axis = "subject_id", x_axis_label = "Patient", 
          count_by = c("subject_id", "tissue"),
          fill_type = "isotype", fill_label = "Isotypes", 
          clrs_specific = col_isotype) +
  facet_wrap("tissue", scales = "fixed", nrow = 2) +
  theme(strip.text.x = element_text(size = 12), 
        axis.text.x = element_text(size = 12)) +
    labs(title = "Percent of Isotypes in controls and treated PWH")

# isotype group
plot_pcts(pcts = calc_pcts(data = dplyr::filter(bcr_heavy, !is.na(isotype)),
                           meta_group_by = c("subject_id", "tissue", "HIV_status"),
                           focus_group = "isotype_group"), plot_type = "All", plot_value = "Isotypes",
          x_axis = "subject_id", x_axis_label = "Patient", 
          count_by = c("subject_id", "tissue"),
          fill_type = "isotype_group", fill_label = "Isotypes") +
  facet_wrap("tissue", scales = "fixed", nrow = 2) +
  theme(strip.text.x = element_text(size = 12), 
        axis.text.x = element_text(size = 12)) +
    labs(title = "Percent of Isotypes in controls and treated PWH")
```

#### By cluster

```{r barplot-isotype-gex-annot}
# isotypes and subtypes per cluster
plot_pcts(pcts = calc_pcts(data = dplyr::filter(bcr_heavy, !is.na(isotype)),
                           meta_group_by = c("tissue", "HIV_status", "gex_annotation_final"),
                           focus_group = "isotype"), plot_type = "All", plot_value = "Isotype",
          x_axis = "gex_annotation_final", x_axis_label = "Cell Cluster", 
          count_by = c("tissue", "gex_annotation_final", "HIV_status"),
          fill_type = "isotype", fill_label = "Isotype", 
          clrs_specific = col_isotype) +
  facet_grid(tissue ~ HIV_status, scales = "fixed") +
  theme(strip.text.x = element_text(size = 12), 
        axis.text.x = element_text(size = 12, angle = 45, hjust = 1))

# isotype group per cluster
plot_pcts(pcts = calc_pcts(data = dplyr::filter(bcr_heavy, !is.na(isotype)),
                           meta_group_by = c("tissue", "HIV_status", "gex_annotation_final"),
                           focus_group = "isotype_group"), plot_type = "All", plot_value = "Isotype",
          x_axis = "gex_annotation_final", x_axis_label = "Cell Cluster", 
          count_by = c("tissue", "gex_annotation_final", "HIV_status"),
          fill_type = "isotype_group", fill_label = "Isotype") +
  facet_grid(tissue ~ HIV_status, scales = "fixed") +
  theme(strip.text.x = element_text(size = 12), 
        axis.text.x = element_text(size = 12, angle = 45, hjust = 1))
```

### CSF and blood isotype composition

```{r}
# prepare dataframes
df_isot_gr <- calc_pcts(data = dplyr::filter(bcr_heavy, !is.na(isotype)), 
                   meta_group_by = c("subject_id", "HIV_status", "tissue", "isotype_group"), 
                   focus_group = "isotype_group") %>% 
  dplyr::rename(fraction_iso = Percent) %>%
  group_by(subject_id, HIV_status, isotype_group) %>%
  dplyr::mutate(delta_fraction = fraction_iso[tissue == "CSF"] - fraction_iso[tissue == "PBMC"])

df_isot <- calc_pcts(data = dplyr::filter(bcr_heavy, !is.na(isotype)), 
                   meta_group_by = c("subject_id", "HIV_status", "tissue", "isotype"), 
                   focus_group = "isotype") %>% 
  dplyr::rename(fraction_iso = Percent) %>%
  group_by(subject_id, HIV_status, isotype) %>%
  dplyr::mutate(delta_fraction = fraction_iso[tissue == "CSF"] - fraction_iso[tissue == "PBMC"])
```

```{r}
# 1. Calculate summary statistics by tissue and isotype_group
stats_summary <- df_isot %>%
  dplyr::filter(isotype != "IGHE") %>%
  group_by(tissue, isotype) %>%
  summarize(
    mean_fraction   = mean(fraction_iso, na.rm = TRUE),
    sd_fraction     = sd(fraction_iso, na.rm = TRUE),
    .groups = "drop"
  )

# 2. Perform Wilcoxon test for each isotype_group comparing fraction_iso between tissues
wilcox_results <- df_isot %>%
  dplyr::filter(isotype != "IgE") %>%
  group_by(isotype) %>%
  do(tidy(wilcox.test(fraction_iso ~ tissue, data = .))) %>%
  ungroup() %>%
  select(isotype, p.value) %>%
  rename(p_value = p.value) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH"))

# 3. Combine the summary statistics for each tissue into one table (wide format)
#    and merge the Wilcoxon test results.
stats_summary %>%
  pivot_wider(names_from = tissue, 
              values_from = c(mean_fraction, sd_fraction, median_fraction, Q1, Q3)) %>%
  left_join(wilcox_results, by = "isotype")
```

#### Isotype comparisons between PWH and PWoH

```{r}
# compute p-value between PWH and PWoH
df_isot_gr %>%
  group_by(isotype_group, tissue) %>%
  summarise(p_value = if (all(table(HIV_status) > 1)) wilcox.test(fraction_iso ~ HIV_status)$p.value else NA_real_, .groups = "drop") %>%
  mutate(p_adj_BH = p.adjust(p_value, method = "BH")) %>% 
  dplyr::filter(p_value < 0.05)

df_isot %>%
  group_by(isotype, tissue) %>%
  summarise(p_value = if (all(table(HIV_status) > 1)) wilcox.test(fraction_iso ~ HIV_status)$p.value else NA_real_, .groups = "drop") %>%
  mutate(p_adj_BH = p.adjust(p_value, method = "BH")) %>% 
  dplyr::filter(p_adj_BH < 0.05)

# remove objects
rm(df_isot_gr, df_isot)
```

#### Isotype by cluster per HIV status and tissue

```{r}
# prepare dataframe for plotting
df_isot <- calc_pcts(data = dplyr::filter(bcr_heavy, !is.na(isotype)), 
                   meta_group_by = c("subject_id", "HIV_status", "tissue", "gex_annotation_final", "isotype_group"), 
                   focus_group = "isotype_group") %>% 
  dplyr::rename(fraction_iso = Percent) %>%
  dplyr::filter(!isotype_group %in% c("IgE") & !gex_annotation_final %in% c("MBC class-unswitched"))

# subset to include cell types with sufficient cell numbers
df_isot_blood <- df_isot %>% dplyr::filter(tissue == "PBMC" & gex_annotation_final != "MBC IgD+")
df_isot_CSF <- df_isot %>% dplyr::filter(tissue == "CSF" & gex_annotation_final == "MBC class-switched" & isotype_group %in% c("IgG", "IgA"))

# p-value correction for multiple testing
df_isot_blood %>%
  group_by(isotype_group, gex_annotation_final, tissue) %>%
  summarise(p_value = if (all(table(HIV_status) > 1)) wilcox.test(fraction_iso ~ HIV_status)$p.value else NA_real_, .groups = "drop") %>%
  mutate(p_adj_BH = p.adjust(p_value, method = "BH")) %>% 
  dplyr::filter(p_adj_BH < 0.05) %>%
  arrange(gex_annotation_final)

# remove objects
rm(df_isot, df_isot_blood, df_isot_CSF)
```

## SHM analysis

### Distribution of SHM across cell clusters

```{r}
# calculate the median mutation frequency of a clone
mut_freq_clone_PWH <- bcr_heavy %>%
                    dplyr::filter(gex_annotation_final != "Naive") %>%
                    dplyr::group_by(subject_id, clone_id_unique, isotype, gex_annotation_final, tissue, HIV_status) %>%
                    dplyr::summarize(mean_mu_freq = mean(mu_freq))

# plot mu_freq histogram
ggplot(mut_freq_clone_PWH, aes(mean_mu_freq)) +
  geom_histogram(binwidth = 0.005) +
  facet_grid(tissue + HIV_status ~ gex_annotation_final, scales = "free_y") +
  labs(x = "Average SHM Frequency", y = "Number of Clonotypes") +
  theme_bw() + theme(plot.title = element_text(size = 18, hjust = 0.5))
```

### Comparison by cluster

```{r}
# check distribution of mu_freq within each individual
ggplot(bcr_heavy %>% dplyr::filter(!gex_annotation_final %in% c("Naive", "MBC_IGD")), 
       aes(y = mu_freq, x = subject_id, fill = HIV_status)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.3) +
  scale_fill_manual(values = col_HIV_status) +
  scale_y_continuous(labels = function(x) paste0(scales::percent(x, accuracy = 1)), breaks = seq(0, 0.3, 0.025)) +
  labs(y = "Mutation frequency", fill = "subject_id",
       title = "Mutation frequency in treated PWH") +
  theme_bw +
  theme(plot.title = element_text(size = 18, hjust = 0.5)) +
  facet_grid(tissue ~ gex_annotation_final, scales = "free") +
  NoLegend()
```

```{r}
# prepare dataframe
data_mu_freq_clust_PWH_ctrl <- bcr_heavy %>% 
                                    dplyr::filter(gex_annotation_final %in% c("MBC class-unswitched", "MBC class-switched")) %>%
                                    group_by(subject_id, HIV_status, tissue, gex_annotation_final) %>%
                                    dplyr::summarize(median_mu_freq = median(mu_freq), .groups = "drop")

# Compute p-values by group (ignoring HIV_status)
pvalues <- data_mu_freq_clust_PWH_ctrl %>%
  group_by(gex_annotation_final, tissue) %>%
  summarise(
    p_value = if (all(table(HIV_status) > 1)) wilcox.test(median_mu_freq ~ HIV_status)$p.value else NA_real_,
    .groups = "drop") 

# Compute median, Q1, and Q3 for each group by HIV_status
summaries <- data_mu_freq_clust_PWH_ctrl %>% 
  dplyr::filter(gex_annotation_final %in% c("MBC class-unswitched", "MBC class-switched")) %>%
  group_by(gex_annotation_final, tissue, HIV_status) %>%
  summarise(
    median_mu_freq_median = median(median_mu_freq),
    Q1 = quantile(median_mu_freq, 0.25),
    Q3 = quantile(median_mu_freq, 0.75),
    .groups = "drop")

# Join the two tables by gex_annotation_final and tissue
combined_table <- left_join(summaries, pvalues, by = c("gex_annotation_final", "tissue"))
combined_table
```

```{r}
# prepare table
delta_tissue_mu_freq <- bcr_heavy %>%
                             dplyr::filter(gex_annotation_final %in% c("MBC class-unswitched", "MBC class-switched")) %>%
                             group_by(subject_id, HIV_status, tissue, gex_annotation_final) %>%
                             dplyr::summarize(median_mu_freq = median(mu_freq), .groups = "drop") %>%
                             group_by(subject_id, HIV_status, gex_annotation_final) %>%
                             dplyr::reframe(tissue_delta_median = median_mu_freq[tissue == "CSF"] - median_mu_freq[tissue == "PBMC"])

# Compute p-values by group (ignoring HIV_status)
pvalues <- delta_tissue_mu_freq %>%
  group_by(gex_annotation_final) %>%
  summarise(
    p_value = if (all(table(HIV_status) > 1)) wilcox.test(tissue_delta_median ~ HIV_status)$p.value else NA_real_,
    .groups = "drop") 

# Compute median, Q1, and Q3 for each group by HIV_status
summaries <- delta_tissue_mu_freq %>% 
  group_by(gex_annotation_final, HIV_status) %>%
  summarise(
    median_mu_freq_median = median(tissue_delta_median),
    Q1 = quantile(tissue_delta_median, 0.25),
    Q3 = quantile(tissue_delta_median, 0.75),
    .groups = "drop")

# Join the two tables by gex_annotation_final and tissue
combined_table <- left_join(summaries, pvalues, by = c("gex_annotation_final"))
combined_table
```

### Comparison by cluster and isotype

```{r}
# prepare table
data_mu_freq_clust_iso_PWH_ctrl <- bcr_heavy %>% dplyr::filter(isotype_group != "IgE") %>%
                                    dplyr::filter(gex_annotation_final %in% c("MBC class-switched")) %>%
                                    group_by(subject_id, HIV_status, tissue, gex_annotation_final, isotype_group) %>%
                                    dplyr::summarize(median_mu_freq = median(mu_freq), .groups = "drop")

# Compute p-values by group (ignoring HIV_status)
pvalues <- data_mu_freq_clust_iso_PWH_ctrl %>%
  group_by(gex_annotation_final, tissue, isotype_group) %>%
  summarise(
    p_value = if (all(table(HIV_status) > 1)) wilcox.test(median_mu_freq ~ HIV_status)$p.value else NA_real_,
    .groups = "drop") 

# Compute median, Q1, and Q3 for each group by HIV_status
summaries <- data_mu_freq_clust_iso_PWH_ctrl %>% 
  group_by(gex_annotation_final, HIV_status, tissue, isotype_group) %>%
  summarise(
    median_mu_freq_median = median(median_mu_freq),
    Q1 = quantile(median_mu_freq, 0.25),
    Q3 = quantile(median_mu_freq, 0.75),
    .groups = "drop")

# Join the two tables by gex_annotation_final and tissue
combined_table <- left_join(summaries, pvalues, by = c("gex_annotation_final", "tissue", "isotype_group"))
combined_table
```

```{r}
# treated PWH and controls
data_mu_freq_clust_iso_PWH_ctrl <- bcr_heavy %>% 
                                    dplyr::filter(gex_annotation_final %in% c("MBC class-switched")) %>%
                                    group_by(subject_id, HIV_status, tissue, gex_annotation_final, isotype_group) %>%
                                    dplyr::summarize(median_mu_freq = median(mu_freq), .groups = "drop") %>%
                                    group_by(isotype_group, gex_annotation_final, HIV_status, subject_id) %>%
                                    dplyr::reframe(delta_median_mu_freq = median_mu_freq[tissue == "CSF"] - median_mu_freq[tissue == "PBMC"])

# Compute p-values by group (ignoring HIV_status)
pvalues <- data_mu_freq_clust_iso_PWH_ctrl %>%
  group_by(gex_annotation_final, isotype_group) %>%
  summarise(
    p_value = if (all(table(HIV_status) > 1)) wilcox.test(delta_median_mu_freq ~ HIV_status)$p.value else NA_real_,
    .groups = "drop") 

# Compute median, Q1, and Q3 for each group by HIV_status
summaries <- data_mu_freq_clust_iso_PWH_ctrl %>% 
  group_by(gex_annotation_final, HIV_status, isotype_group) %>%
  summarise(
    median_mu_freq_median = median(delta_median_mu_freq),
    Q1 = quantile(delta_median_mu_freq, 0.25),
    Q3 = quantile(delta_median_mu_freq, 0.75),
    .groups = "drop")

# Join the two tables by gex_annotation_final and tissue
combined_table <- left_join(summaries, pvalues, by = c("gex_annotation_final", "isotype_group"))
combined_table
```

```{r}
# remove objects
rm(data_mu_freq_clust_iso_PWH_ctrl, data_mu_freq_clust_PWH_ctrl, data_mu_freq_iso_PWH_ctrl, delta_tissue_mu_freq)
```

## Epitope Specificity

```{r}
# load IEDB epitotope sequences for HIV downloaded here: https://www.iedb.org/
aa_seq_IEDB_heavy <- read.csv(header = TRUE, "./receptor_table_export_1730925113.csv") %>% #   HIV_epitopes_IEDB.csv
  dplyr::select(organism, molecule, chain1_vgene, chain1_jgene, chain1_cdr3_aa, chain2_vgene, chain2_jgene, chain2_cdr3_aa) %>% 
  mutate(chain1_vgene = getGene(chain1_vgene, first = FALSE, collapse = TRUE, strip_d = TRUE),
         chain1_jgene = getGene(chain1_jgene, first = FALSE, collapse = TRUE, strip_d = TRUE))

# subset data to MBC heavy chains and separate ambiguous v and j calls
seq_to_check <- bcr_data %>% 
  dplyr::filter(locus == "IGH") %>%
  #dplyr::filter(gex_annotation_final %in% c("MBC class-unswitched", "MBC class-switched")) %>% 
  dplyr::select(cell_id_unique, v_call, j_call, junction_aa, locus) %>%
  separate_rows(v_call, sep = ",") %>% 
  separate_rows(j_call, sep = ",") %>% 
  mutate(junction_aa = substr(junction_aa, 2, nchar(junction_aa) - 1)) %>%
  dplyr::rename(cdr3_aa = junction_aa) %>%
  mutate(v_call = getGene(v_call, first = FALSE, collapse = TRUE, strip_d = TRUE),
         j_call = getGene(j_call, first = FALSE, collapse = TRUE, strip_d = TRUE))

# match data with IEDB epitopes
seq_to_check_VJ_match <- seq_to_check %>% 
  inner_join(aa_seq_IEDB_heavy, by = join_by("v_call" == "chain1_vgene", "j_call" == "chain1_jgene"), relationship = "many-to-many") %>% 
  distinct() %>%
  dplyr::filter(nchar(cdr3_aa) == nchar(chain1_cdr3_aa))

# Calculate sequence distance for each row
seq_to_check_VJ_match <- seq_to_check_VJ_match %>%
  rowwise() %>% # apply function row-wise
  mutate(seq_distance = seqDist(cdr3_aa, chain1_cdr3_aa, dist_mat = getAAMatrix()),
         seq_distance_norm = seq_distance/nchar(cdr3_aa)) %>% # calculate distance
  ungroup()

# find sequence with low distance
seq_match_select <- seq_to_check_VJ_match %>% dplyr::filter(seq_distance < 3)

# check for light chain similarity
seq_to_check <- bcr_data %>% 
  dplyr::filter(locus != "IGH" & cell_id_unique %in% seq_match_select$cell_id_unique) %>%
  dplyr::select(cell_id_unique, v_call, j_call, junction_aa, locus) %>%
  separate_rows(v_call, sep = ",") %>% 
  separate_rows(j_call, sep = ",") %>% 
  mutate(junction_aa = substr(junction_aa, 2, nchar(junction_aa) - 1)) %>%
  dplyr::rename(cdr3_aa = junction_aa) %>%
  mutate(v_call = getGene(v_call, first = FALSE, collapse = TRUE, strip_d = TRUE),
         j_call = getGene(j_call, first = FALSE, collapse = TRUE, strip_d = TRUE))

seq_to_check_VJ_match_light <- seq_match_select %>%
  mutate(chain2_vgene = getGene(chain2_vgene, first = FALSE, collapse = TRUE, strip_d = TRUE),
         chain2_jgene = getGene(chain2_jgene, first = FALSE, collapse = TRUE, strip_d = TRUE)) %>%
  inner_join(aa_seq_IEDB_heavy, by = join_by("v_call" == "chain2_vgene", "j_call" == "chain2_jgene"), relationship = "many-to-many") %>% 
  distinct() %>%
  dplyr::filter(nchar(cdr3_aa) == nchar(chain2_cdr3_aa))

# remove objects
rm(aa_seq_IEDB_heavy, seq_to_check, seq_to_check_VJ_match, seq_match_select, seq_to_check_VJ_match_light)
```
