## Packages

```{r}
# load packages
packages <- c("tidyverse", "scales", "rlang", "magrittr", "ggpubr",               # data tidying and base packages
              "ggplot2", "patchwork", "pheatmap", "ggVennDiagram",                # visualization and plotting
              "Seurat", "clustree", "scDblFinder", "Azimuth", "clusterProfiler",  # gene expression analysis
              "airr", "alakazam", "shazam", "scoper", "dowser")                   # airr analysis
for (n in 1:length(packages)) {
    suppressPackageStartupMessages(library(packages[n], character.only = TRUE))
    cat(paste0(packages[n], ": ", packageVersion(packages[n]), "\n"))
}

# remove objects
rm(n, packages)
```

## Paths

```{r}
# assign important paths
path_objs <- file.path("..", "objects")
path_data_GEX <- file.path("data", "cellranger", "gex")
path_ref_genes <- file.path("..", "reference_genes")
path_airr_BCR <- file.path("..", "data", "airrflow", "bcr", "clonal_analysis", "define_clones")
```

## Colors and themes

```{r colors}
col_tissue <- c("PBMC" = "#fcbba1", "CSF" = "#c6dbef")
col_choroid <- c("darkgreen")
col_clustree <- setNames(pals::stepped2(n = 20), nm = seq(0.1, 1.6, 0.1))
col_isotype <- c("IGHD" = "#ef6548", "IGHM" = "#fec44f", "IGHE" = "#810f7c", "NA" = "lightgrey",
                  "IGHG1" = "#08519c", "IGHG2" = "#3182bd", "IGHG3" = "#6baed6", "IGHG4" = "#bdd7e7",
                  "IGHA1" = "#31a354", "IGHA2" = "#a1d99b")
col_isotype_group <- c("IgD" = "red", "IgM" = "#fe9929", "IgE" = "#810f7c", "NA" = "lightgrey",
                  "IgG" = "darkblue", "IgA" = "darkgreen")
col_locus <- c("IGH" = "#8856a7", "IGK" = "#a6bddb", "IGL" = "#a8ddb5")
col_HIV_status <- c("control" = "#c7e9c0", "PWH" = "#bcbddc")
col_HIV_status_PCA <- c("control" = "#41ab5d", "PWH" = "#9e9ac8")

# define colors for each individual
extended_palette <- colorRampPalette(RColorBrewer::brewer.pal(12, "Set3"))
col_subj <- extended_palette(18)
values <- unique(gex_valid$subject_id)
names(col_subj) <- values[1:length(col_subj)]
rm(extended_palette, values)
```

```{r}
theme_bw <- theme_bw() + 
              theme(panel.grid.major.x = element_blank(),
                    panel.grid.major.y = element_line(color = "grey75",
                                                      linewidth = 0.2))

labels_rotate_x <- theme(axis.text.x = element_text(angle = 45, hjust = 1))

labels_standard <- theme(plot.title = element_text(size = 15, hjust = 0.5),
                         plot.subtitle = element_text(size = 12, hjust = 0.5),
                         plot.caption = element_text(size = 7),
                         axis.title = element_text(size = 9),
                         axis.text = element_text(size = 8),
                         legend.title = element_text(size = 10),
                         legend.text = element_text(size = 8))

clean_umap <- theme(plot.title = element_text(hjust = 0.5),
                    axis.text = element_blank(),
                    axis.ticks = element_blank())
```

## Functions
## Metadata summarization

```{r}
summarize_meta <- function(df, 
                           remove_vars = NULL, 
                           group_var,                 
                           cont_method = "wilcox",    # "wilcox" (default) or "t.test"
                           cat_method = "dynamic",    # "dynamic" (default), "chisq", or "fisher"
                           chi_correct = TRUE         # For chi-square: enable (TRUE) or disable (FALSE) Yate's continuity correction
                           ) {
  # Remove unwanted variables if provided
  if (!is.null(remove_vars)) {
    df <- df %>% dplyr::select(-all_of(remove_vars))
  }
  
  # Check that the grouping variable exists
  if (!group_var %in% names(df)) {
    stop("The grouping variable ", group_var, " is not found in the dataframe.")
  }
  
  ### Helper functions ###
  # Continuous variable summary: calculates median (Q1–Q3) per group.
  summarize_continuous <- function(data, var, group_var) {
    data %>%
      group_by(across(all_of(group_var))) %>%
      summarise(
        Median = median(.data[[var]], na.rm = TRUE),
        Q1 = quantile(.data[[var]], 0.25, na.rm = TRUE),
        Q3 = quantile(.data[[var]], 0.75, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(Statistic = paste0(Median, " (", Q1, "-", Q3, ")")) %>%
      dplyr::select(all_of(group_var), Statistic)
  }
  
  # Categorical variable summary: excludes NAs and creates a "Label" column (the level).
  categorical_summary <- function(data, var, group_var) {
    data %>%
      filter(!is.na(.data[[var]])) %>%  # Exclude NA values
      group_by(Label = .data[[var]], across(all_of(group_var))) %>%
      summarise(n = n(), .groups = "drop") %>%
      group_by(across(all_of(group_var))) %>%
      mutate(perc = round(100 * n / sum(n), 1)) %>%
      ungroup() %>%
      mutate(Percentage = paste0(n, " (", perc, "%)")) %>%
      dplyr::select(Label, all_of(group_var), Percentage)
  }
  
  ### Main Loop ###
  # Get all variables except the grouping variable
  vars_to_summarize <- setdiff(names(df), group_var)
  
  # Initialize a list to collect summary outputs
  summary_list <- list()
  
  for (var in vars_to_summarize) {
    if (is.numeric(df[[var]])) {
      # --- Continuous variable ---
      cont_sum <- summarize_continuous(df, var, group_var) %>%
        pivot_wider(names_from = all_of(group_var), values_from = Statistic)
      
      # Prepare data for statistical test (exclude missing values)
      tmp <- df %>%
        dplyr::select(all_of(c(var, group_var))) %>%
        filter(!is.na(.data[[var]]), !is.na(.data[[group_var]]))
      
      # Check that exactly two groups are present and each has at least one observation
      if (length(unique(tmp[[group_var]])) == 2 && all(table(tmp[[group_var]]) > 0)) {
        if (cont_method == "wilcox") {
          test_result <- tryCatch(
            wilcox.test(as.formula(paste(var, "~", group_var)), data = tmp),
            error = function(e) NULL
          )
          test_name <- "Wilcoxon rank-sum"
        } else if (cont_method == "t.test") {
          test_result <- tryCatch(
            t.test(as.formula(paste(var, "~", group_var)), data = tmp),
            error = function(e) NULL
          )
          test_name <- "t-test"
        } else {
          test_result <- NULL
          test_name <- NA
        }
        p_val <- if (!is.null(test_result)) test_result$p.value else NA
      } else {
        p_val <- NA
        test_name <- NA
      }
      
      # If no p-value was calculated, then remove the test name.
      if (is.na(p_val)) {
        test_name <- NA
      }
      
      cont_sum <- cont_sum %>%
        mutate(Variable = var,
               Label = NA,
               Type = "Continuous",
               p_value = p_val,
               Test = test_name) %>%
        dplyr::select(Variable, Label, Type, everything())
      
      summary_list[[var]] <- cont_sum
      
    } else {
      # --- Categorical variable ---
      cat_sum <- categorical_summary(df, var, group_var) %>%
        pivot_wider(names_from = all_of(group_var), values_from = Percentage)
      
      # Build contingency table using non-NA rows for both the variable and group.
      df_subset <- df %>% filter(!is.na(.data[[var]]), !is.na(.data[[group_var]]))
      tab <- table(df_subset[[var]], df_subset[[group_var]])
      
      # Check that each group (each column in the table) has at least one observation.
      if (any(colSums(tab) == 0)) {
        p_val <- NA
        test_name <- NA
      } else {
        if (cat_method == "dynamic") {
          # Use chi-square if every cell has at least 5; otherwise, use Fisher's exact.
          if (all(tab >= 5)) {
            test_result <- tryCatch(chisq.test(tab, correct = chi_correct), error = function(e) NULL)
            test_name <- "Chi-square"
          } else {
            test_result <- tryCatch(fisher.test(tab), error = function(e) NULL)
            test_name <- "Fisher's exact"
          }
        } else if (cat_method == "chisq") {
          test_result <- tryCatch(chisq.test(tab, correct = chi_correct), error = function(e) NULL)
          test_name <- "Chi-square"
        } else if (cat_method == "fisher") {
          test_result <- tryCatch(fisher.test(tab), error = function(e) NULL)
          test_name <- "Fisher's exact"
        } else {
          test_result <- NULL
          test_name <- NA
        }
        p_val <- if (!is.null(test_result)) test_result$p.value else NA
      }
      
      if (is.na(p_val)) {
        test_name <- NA
      }
      
      cat_sum <- cat_sum %>%
        mutate(Variable = var,
               Type = "Categorical",
               p_value = p_val,
               Test = test_name) %>%
        dplyr::select(Variable, Label, Type, everything())
      
      summary_list[[var]] <- cat_sum
    }
  }
  
  # Combine all summaries into one final table and arrange the rows.
  summary_table <- bind_rows(summary_list) %>%
    arrange(Type, Variable, Label)
  
  return(summary_table)
}
```


## Quality control

### Metrics summary for cellranger count output

```{r}
create_metrics_summary_count <- function(meta, path_data_specific,
                                   data_types, samples_list) {
  metrics_summary <- tibble()
  if (missing(data_types)) data_types <- unique(meta$data_type)
  if (missing(samples_list)) samples_list <- sort(unique(meta$sample_name))

  for (sample in samples_list) {
    for (data_type in data_types) {
      # metadata info
      meta_sample <- dplyr::filter(meta, sample_name == sample, data_type == data_type)
      subject_id <- unique(meta_sample$subject_id)
      tissue <- unique(meta_sample$tissue)

      # read in the summary table
      data <- read_csv(file.path(path_data_specific, meta_sample$sample_name,
                                 "outs", 
                                 "metrics_summary.csv"),
                       show_col_types = FALSE)

      # set the columns of interest
      estimated_number_of_cells <- data[["Estimated Number of Cells"]]
      num_reads <- data[["Number of Reads"]] # Number of Read Pairs (VDJ), Number of Reads (GEX)
      reads_per_cell <- data[["Mean Reads per Cell"]]
      median_genes <- data[["Median Genes per Cell"]] # GEX only
      total_genes <- data[["Total Genes Detected"]]
      valid_barcodes <- data[["Valid Barcodes"]]
      UMI_counts_cell <- data[["Median UMI Counts per Cell"]]
      reads_in_cells <- data[["Fraction Reads in Cells"]]
      Q30_bases_in_read <- data[["Q30 Bases in RNA Read"]]
      seq_saturation <- data[["Sequencing Saturation"]]
      intronic <- data[["Reads Mapped Confidently to Intronic Regions"]]
      exonic <- data[["Reads Mapped Confidently to Exonic Regions"]]
      transcriptome_map <- data[["Reads Mapped Confidently to Transcriptome"]]
      antisense <- data[["Reads Mapped Antisense to Gene"]]
      
      # update the combined summary table
      metrics_summary <- bind_rows(metrics_summary,
                                   tibble(sample, subject_id, tissue, estimated_number_of_cells,
                                          num_reads, reads_per_cell, reads_in_cells, median_genes,
                                          total_genes, valid_barcodes, UMI_counts_cell, seq_saturation, 
                                          Q30_bases_in_read, intronic, exonic, transcriptome_map, 
                                          antisense, data_type))
    }
  }

  colnames(metrics_summary) <- c("sample_name", "subject_id", "tissue", "EstimatedNumberofCells",
                                 "NumberofReads", "MeanReadsPerCell", "ReadsInCells", "MedianGenesPerCell",
                                 "TotalGenes", "ValidBarcodes", "MedianUMIPerCell", "SeqSaturation", 
                                 "Q30basesInRead", "IntronicRegions", "ExonicRegions", "TranscriptomeMapping", 
                                 "AntisenseMapping", "data_type")
  return(metrics_summary)
}
```

### Metrics summary for cellranger ARC output

```{r}
create_metrics_summary_arc <- function(meta, path_data_specific,
                                   data_types, samples_list) {
  metrics_summary <- tibble()
  if (missing(data_types)) data_types <- unique(meta$data_type)
  if (missing(samples_list)) samples_list <- sort(unique(meta$sample_name))

  for (sample in samples_list) {
    for (data_type in data_types) {
      # metadata info
      meta_sample <- dplyr::filter(meta, sample_name == sample, data_type == data_type)
      subject_id <- unique(meta_sample$subject_id)
      tissue <- unique(meta_sample$tissue)

      # read in the summary table
      data <- read_csv(file.path(path_data_specific, meta_sample$sample_name,
                                 "summary.csv"),
                       show_col_types = FALSE)

      # set the columns of interest
      estimated_number_of_cells <- data[["Estimated number of cells"]]
      num_reads <- data[["GEX Sequenced read pairs"]] # Number of Read Pairs (VDJ), Number of Reads (GEX)
      reads_per_cell <- data[["GEX Mean raw reads per cell"]]
      median_genes <- data[["GEX Median genes per cell"]] 
      total_genes <- data[["GEX Total genes detected"]] 
      valid_barcodes <- data[["GEX Valid barcodes"]]
      valid_UMI <- data[["GEX Valid UMIs"]] 
      UMI_counts_cell <- data[["GEX Median UMI counts per cell"]] 
      reads_in_cells <- data[["GEX Fraction of transcriptomic reads in cells"]]
      Q30_bases_in_read_2 <- data[["GEX Q30 bases in read 2"]]
      exonic <- data[["GEX Reads mapped confidently to exonic regions"]]
      intronic <- data[["GEX Reads mapped confidently to intronic regions"]]
      transcriptome_map <- data[["GEX Reads mapped confidently to transcriptome"]]
      antisense <- data[["GEX Reads mapped antisense to gene"]]
      
      # update the combined summary table
      metrics_summary <- bind_rows(metrics_summary,
                                   tibble(sample, subject_id, tissue, estimated_number_of_cells,
                                          num_reads, reads_per_cell, reads_in_cells, median_genes,
                                          total_genes, valid_barcodes, valid_UMI, UMI_counts_cell, 
                                          Q30_bases_in_read_2, intronic, exonic, transcriptome_map, 
                                          antisense, data_type))
    }
  }

  colnames(metrics_summary) <- c("sample_name", "subject_id", "tissue", "EstimatedNumberofCells",
                                 "NumberofReads", "MeanReadsPerCell", "ReadsInCells", "MedianGenesPerCell",
                                 "TotalGenes", "ValidBarcodes", "ValidUMI", "MedianUMIPerCell", 
                                 "Q30basesRead2", "IntronicRegions", "ExonicRegions", "TranscriptomeMapping", 
                                 "AntisenseMapping", "data_type")
  return(metrics_summary)
}
```


## General plots

### Bar plots with stacked percentages

```{r}
calc_pcts <- function(data, meta_group_by = c("sample_name", "batch"),
                      focus_group) {
  data %>%
    group_by(across(all_of(meta_group_by)), !!sym(focus_group)) %>%
    summarize(Count = n(), .groups = "drop_last") %>%
    dplyr::rename(focus = !!sym(focus_group)) %>%
    complete(focus = unique(data[[focus_group]]), fill = list(Count = 0)) %>%
    mutate(Percent = 100 * Count / sum(Count)) %>%
    dplyr::rename(!!sym(focus_group) := focus) %>%
    ungroup()
}
```

```{r}
plot_pcts <- function(pcts, tissue_type, clrs_specific,
                      plot_type = "All", plot_value = "Cell Type",
                      x_axis = "sample_name", x_axis_label = "Sample",
                      count_by = c("sample_name", "tissue"),
                      fill_type = "annotated_clusters", fill_label = fill_type,
                      perc_min = 3, label_size = 3, label_fill = FALSE,
                      drop_zeroes = TRUE, details = NULL, draw_border = TRUE,
                      remove_pct = FALSE, remove_count = FALSE) {
  # if you want to use the default Seurat colors
  if (is_missing(clrs_specific))
    clrs_specific <- hue_pal()(n_distinct(pcts %>%
                                            ungroup() %>%
                                            dplyr::select(all_of(fill_type))))

  # mainly affects the legend
  if (drop_zeroes) pcts <- dplyr::filter(pcts, Percent != 0)

  # set up the plotting layers properly
  if (plot_type == "All") {
    # add the total counts to the data
    pcts <- pcts %>%
              group_by(across(all_of(count_by))) %>%
              mutate(Total = sum(Count), Total = replace(Total, -n(), ""))
  }
  else if (plot_type == "Binary") {
    # so that TRUE plots on top of FALSE
    pcts[[fill_type]] <- factor(pcts[[fill_type]], levels = c(TRUE, FALSE))
  }
  else {
    stop("Please enter a valid plot type.")
  }

  # set up the plot with conditional border drawing
  p <- ggplot(data = pcts,
              aes(x = !!sym(x_axis), y = Percent, fill = !!sym(fill_type))) +
        geom_bar(stat = "identity", 
                 color = if (draw_border) "black" else NA,
                 linewidth = if (draw_border) 0.2 else 0) +
        labs(title = paste("Percent of", plot_value, "per", x_axis_label),
             subtitle = details,
             x = x_axis_label, y = paste(plot_value, "Percentage"),
             fill = fill_label)

  # add aesthetics
  p <- p + scale_fill_manual(values = clrs_specific, limits = force) +
           scale_y_continuous(labels = label_percent(scale = 1),
                              expand = expansion(mult = 0.05))

  # add percentage labels only if remove_pct is FALSE
  if (!remove_pct) {
    if (!label_fill) {
      p <- p + geom_text(aes(label = label_percent(accuracy = 1, scale = 1)(
                                 ifelse(Percent > perc_min, Percent, NA)),
                             group = !!sym(fill_type)),
                         size = label_size,
                         position = position_stack(vjust = 0.5))
    }
    else {
      p <- p + geom_label(aes(label = ifelse(Percent > perc_min,
                                             percent(Percent, accuracy = 1, scale = 1),
                                             NA),
                              group = !!sym(fill_type)),
                          color = "black", fill = "white",
                          size = label_size,
                          position = position_stack(vjust = 0.5))
    }
  }

  # add count text only if remove_count is FALSE
  if (!remove_count) {
    if (plot_type == "All") {
      p <- p + geom_text(mapping = aes(x = !!sym(x_axis), y = 100,
                                       label = Total, fill = NULL),
                         vjust = -0.5, color = "black", size = label_size)
    }
    else if (plot_type == "Binary") {
      p <- p +
            # counts on top
            geom_text(mapping = aes(label = ifelse(!!sym(fill_type) == TRUE,
                                                   Count, "")),
                      position = position_stack(), vjust = -0.5,
                      color = "black", size = label_size) +
            # counts on bottom
            geom_text(mapping = aes(label = ifelse(!!sym(fill_type) == FALSE,
                                                   Count, ""), y = -1),
                      color = "black", size = label_size)
    }
  }

  # add remaining styling
  p <- p + theme_bw + labels_standard

  # only rotate long names
  if (x_axis %in% c("sample_name",
                    "annotated_clusters", "annotated_subclusters")) {
    p <- p + labels_rotate_x
  }

  return(p)
}
```

### Top info bar

```{r}
add_info_bar <- function(plot, method = "contains", info_type, info,
                         text_size = 8, label = TRUE) {
  # add the information to the plot data if needed
  if (method == "add") {
    plot$data[info_type] <- info[[info_type]]
  }
  else if (method == "join") {
    plot$data <- suppressMessages(left_join(plot$data, info, # by = {{info_type}},
                                            multiple = "all",
                                            relationship = "many-to-many"))
  }

  # explain the label or not
  if (label) labels <- purrr::partial(label_both, sep = " ")
  else labels <- labeller()

  # split up the plot
  plot + facet_grid(cols = vars(!!sym(info_type)),
                    scales = "free_x", space = "free", labeller = labels) +
    theme(strip.text.x = element_text(size = text_size))
}
```

### Dot plot with side info

```{r}
plot_dot_side <- function(plot, seurat_obj, row_identity = "seurat_clusters",
                          facet_col, info_to_add = c("cluster_size")) {
  # get the clusters in order
  row_idents <- levels(plot$data$id)
  side_labels <- data.frame("row_idents" = row_idents)

  # calculate extra useful information
  if ("cluster_size" %in% info_to_add) {
    counts_sum <- seurat_obj[[]] %>%
                    dplyr::count(!!sym(row_identity)) %>%
                    arrange(factor(!!sym(row_identity), levels = row_idents)) %>%
                    pull(n)

    # reformat
    counts_sum <- str_pad(string = as.character(counts_sum),
                          width = max(nchar(counts_sum)),
                          side = "left", pad = " ")

    side_labels <- bind_cols(side_labels, "total counts" = counts_sum)
  }

  if ("mean_mu_freq" %in% info_to_add) {
    mean_mu_freq <- seurat_obj[[]] %>%
                      group_by(!!sym(row_identity)) %>%
                      arrange(factor(!!sym(row_identity), levels = row_idents)) %>%
                      mutate(mean_mu_freq = mean(mu_freq, na.rm = TRUE)) %>%
                      distinct(mean_mu_freq) %>%
                      pull()

    # reformat
    # mean_mu_freq <- sprintf("%.2f", mean_mu_freq)
    mean_mu_freq <- label_percent(accuracy = 0.1)(mean_mu_freq)
    mean_mu_freq <- replace_na(mean_mu_freq, replace = "NA")
    mean_mu_freq <- str_pad(string = as.character(mean_mu_freq),
                          width = max(nchar(mean_mu_freq)),
                          side = "left", pad = " ")

    side_labels <- bind_cols(side_labels, "mean mutation frequency" = mean_mu_freq)
  }

  if ("percent_BCR" %in% info_to_add) {
    # needs more work
    percent_BCR <- calc_pcts(data = seurat_obj[[]],
                             meta_group_by = row_identity,
                             focus_group = "contains_bcr") %>%
                    dplyr::filter(contains_bcr) %>%
                    # mutate(Percent = round(Percent, digits = 1)) %>%
                    arrange(factor(!!sym(row_identity), levels = row_idents)) %>%
                    pull(Percent)

    # reformat
    percent_BCR <- label_percent(accuracy = 1, scale = 1)(percent_BCR)
    percent_BCR <- str_pad(string = as.character(percent_BCR),
                           width = max(nchar(percent_BCR)),
                           side = "left", pad = " ")

    side_labels <- bind_cols(side_labels, "percent BCR" = percent_BCR)
  }

  if ("percent_TCR" %in% info_to_add) {
    # needs more work
    percent_TCR <- calc_pcts(data = seurat_obj[[]],
                             meta_group_by = row_identity,
                             focus_group = "contains_tcr") %>%
                    dplyr::filter(contains_tcr) %>%
                    # mutate(Percent = round(Percent, digits = 1)) %>%
                    arrange(factor(!!sym(row_identity), levels = row_idents)) %>%
                    pull(Percent)

    # reformat
    percent_TCR <- label_percent(accuracy = 1, scale = 1)(percent_TCR)
    percent_TCR <- str_pad(string = as.character(percent_TCR),
                           width = max(nchar(percent_TCR)),
                           side = "left", pad = " ")

    side_labels <- bind_cols(side_labels, "percent TCR" = percent_TCR)
  }

  if ("TRUST4" %in% info_to_add) {
    # needs more work
    percent_TRUST4 <- calc_pcts(data = seurat_obj[[]],
                                meta_group_by = row_identity,
                                focus_group = "TRUST4") %>%
      dplyr::filter(TRUST4) %>%
      # mutate(Percent = round(Percent, digits = 1)) %>%
      arrange(factor(!!sym(row_identity), levels = row_idents)) %>%
      pull(Percent)
    
    # reformat
    percent_TRUST4 <- label_percent(accuracy = 1, scale = 1)(percent_TRUST4)
    percent_TRUST4 <- str_pad(string = as.character(percent_TRUST4),
                              width = max(nchar(percent_TRUST4)),
                              side = "left", pad = " ")
    
    side_labels <- bind_cols(side_labels, "percent TRUST4" = percent_TRUST4)
  }

  # add info along the right side
  subtitle_details <- side_labels %>%
                        dplyr::select(-row_idents) %>%
                        colnames() %>%
                        str_c(collapse = " | ")
  side_labels <- tidyr::unite(side_labels, col = "final_labels", -row_idents,
                       sep = " | ") %>%
                   pull(final_labels, name = row_idents)

  if (missing(facet_col)) {
    plot <- plot + facet_grid(rows = vars(id),
                              labeller = labeller(id = side_labels),
                              scales = "free", space = "free")
  }
  else {
    plot <- plot + facet_grid(rows = vars(id), cols = vars(!!sym(facet_col)),
                              labeller = labeller(id = side_labels),
                              scales = "free", space = "free")
  }

  # final plot
  plot +
    labs(subtitle = subtitle_details) +
    theme(strip.text.y = element_text(angle = 0, size = 8),
          strip.background.y = element_rect(fill = "white"),
          plot.subtitle = element_text(size = 9))
}
```

## AIRR analysis in GEX

### Mutation frequency

```{r}
plot_mu_freq <- function(seurat_obj, tissue_type,
                         plot_label = TRUE, label_size = 4) {
  # you need the NAs for the background
  max_freq <- max(seurat_obj$mu_freq, na.rm = TRUE)
  # for a consistent legend scale
  max_freq <- round(max_freq, digits = 2)

  # set up the plot
  p <- FeaturePlot(seurat_obj, features = "mu_freq", pt.size = 0.2,
                   order = TRUE, label = plot_label, label.size = label_size,
                   raster = FALSE) +
         labs(title = paste(tissue_type, "Mutation Frequencies"),
              subtitle = paste0("Max: ", percent(max_freq), ", Mean: ",
                                percent(mean(seurat_obj$mu_freq, na.rm = TRUE)))) +
         scale_color_viridis_c(name = "Mutation Freq",
                               labels = percent, direction = -1,
                               option = "C", na.value = "lightgray",
                               breaks = seq(0, max_freq, 0.05),
                               limits = c(0, max_freq))

  # standardize the labels
  p <- p + labels_standard #+ NoLegend()
  
  return(p)
}
```

### Mutation frequency of isotypes

```{r}
plot_mu_freq_isotype_overlay <- function(seurat_obj, isotypes, tissue_type) {
  # you need the NAs for the background
  obj <- subset(seurat_obj, isotype_group %in% isotypes | is.na(isotype_group))
  max_freq <- max(seurat_obj$mu_freq, na.rm = TRUE) # for a consistent legend scale

  # set up the plot
  p <- FeaturePlot(obj, features = "mu_freq", pt.size = 0.2,
                   order = TRUE, raster = FALSE, repel = TRUE) +
         labs(title = paste0(tissue_type, " (", isotypes, ")"),
              subtitle = paste0("Max: ", percent(max(obj$mu_freq, na.rm = TRUE)),
                                ", Mean: ", percent(mean(obj$mu_freq, na.rm = TRUE)))) +
         scale_color_viridis_c(name = "Mutation Freq",
                               labels = percent, direction = -1,
                               option = "C", na.value = "lightgray",
                               breaks = seq(0, max_freq, 0.05),
                               limits = c(0, max_freq))
  
  # standardize the labels
  p <- p + labels_standard
  
  return(p)
}
```

## Cell Frequencies

```{r}
calculateCellFrequencies <- function(sc_df, 
                                     cluster_col, 
                                     subject_col, 
                                     vars_to_keep = NULL, 
                                     show_clusters = NULL, 
                                     denominator_clusters = NULL) {
  # Load required package
  if (!requireNamespace("dplyr", quietly = TRUE)) {
    stop("The dplyr package is required but not installed. Please install it first.")
  }
  library(dplyr)
  
  # Define grouping variables for counting cells (numerator)
  grouping_vars <- c(subject_col, cluster_col, vars_to_keep)
  
  # Calculate cell counts per subject-cluster (and additional groups if provided)
  cell_counts <- sc_df %>%
    group_by(across(all_of(grouping_vars))) %>%
    summarise(cell_count = n(), .groups = "drop")
  
  # Define grouping variables for the denominator (subject plus any additional variables)
  grouping_vars_total <- c(subject_col, vars_to_keep)
  
  # Calculate total cells per subject (and additional variables) for the denominator.
  # If denominator_clusters is specified, only include those clusters in the total.
  if (!is.null(denominator_clusters)) {
    total_cells <- sc_df %>%
      filter(.data[[cluster_col]] %in% denominator_clusters) %>%
      group_by(across(all_of(grouping_vars_total))) %>%
      summarise(total_cells = n(), .groups = "drop")
  } else {
    total_cells <- sc_df %>%
      group_by(across(all_of(grouping_vars_total))) %>%
      summarise(total_cells = n(), .groups = "drop")
  }
  
  # Merge the cell counts with the total cell counts based on subject and additional variables
  result <- cell_counts %>%
    left_join(total_cells, by = grouping_vars_total) %>%
    mutate(frequency = cell_count / total_cells)
  
  # Apply the show_clusters filter to display only the desired clusters in the final result
  if (!is.null(show_clusters)) {
    result <- result %>%
      filter(.data[[cluster_col]] %in% show_clusters)
  }
  
  # Rename the column specified by cluster_col to 'cluster'
  result <- result %>% dplyr::rename(cluster = !!rlang::sym(cluster_col))
  
  return(result)
}
```

## Pseudobulk analysis

### DGE with DESeq2

```{r}
pseudobulk_DE <- function(seurat_object,
                          replicate,
                          cell_clusters,
                          compare_var,
                          reference_level,
                          covariate = NULL,
                          params = list()) {
  
  ## Check that the required columns exist in the metadata
  required_cols <- c(replicate, cell_clusters, compare_var)
  if (!is.null(covariate)) {
    required_cols <- c(required_cols, covariate)
  }
  missing_cols <- setdiff(required_cols, colnames(seurat_object@meta.data))
  if(length(missing_cols) > 0) {
    stop("The following required columns are missing from seurat_object@meta.data: ", 
         paste(missing_cols, collapse = ", "))
  }
  
  ## Rename metadata columns to standard names used throughout the function
  seurat_object@meta.data$replicate <- as.character(seurat_object@meta.data[[replicate]])
  seurat_object@meta.data$cell_type <- as.character(seurat_object@meta.data[[cell_clusters]])
  seurat_object@meta.data$label     <- as.factor(seurat_object@meta.data[[compare_var]])
  
  ## If a covariate is provided, add it to the metadata
  if(!is.null(covariate)) {
    seurat_object@meta.data$covariate <- seurat_object@meta.data[[covariate]]
  }
  
  ## Relevel the comparison variable factor using the provided reference level
  seurat_object@meta.data$label <- relevel(seurat_object@meta.data$label, ref = reference_level)
  
  ## Save metadata for later use
  meta <- seurat_object@meta.data
  
  ## Check that a replicate column now exists
  if (!"replicate" %in% colnames(meta))
    stop("metadata does not contain replicate information")
  
  ## Keep only cell types with enough cells (at least 3 cells per label per cell type)
  keep <- meta %>%
    dplyr::count(cell_type, label) %>%
    dplyr::group_by(cell_type) %>%
    dplyr::filter(all(n >= 3)) %>%
    dplyr::pull(cell_type) %>%
    unique()
  
  ## Process data into gene x replicate matrices for each cell type
  Idents(seurat_object) <- seurat_object$cell_type
  pseudobulks <- purrr::map(keep, function(ct) {
    message("Processing cell type: ", ct)
    so_sub <- subset(seurat_object, idents = ct)
    
    ## Create a local copy of metadata for this cell type
    meta_sub <- so_sub@meta.data %>%
      dplyr::mutate(label = as.character(label),
                    replicate = as.character(replicate))
    
    ## Check for at least two groups (e.g. at least two levels in 'label')
    if (n_distinct(meta_sub$label) < 2)
      return(NA)
    
    ## Optionally allow per-cell pseudo-bulking if params$replicate == 'cells'
    if (!is.null(params$replicate)) {
      if (params$replicate == 'cells') {
        meta_sub$replicate <- colnames(so_sub)
        ## Remove any ":" characters which may cause issues downstream
        if (any(grepl("\\:", meta_sub$replicate))) {
          meta_sub$replicate <- gsub(".*\\:", "", meta_sub$replicate)
        }
      }
    } else {
      replicate_counts <- dplyr::distinct(meta_sub, label, replicate) %>%
        dplyr::group_by(label) %>%
        dplyr::summarise(replicates = n_distinct(replicate)) %>%
        dplyr::pull(replicates)
      if (any(replicate_counts < 2))
        return(NA)
    }
    
    ## Create a design matrix to aggregate counts per replicate per group.
    ## This uses the combined replicate and label information.
    mm <- model.matrix(~ 0 + replicate:label, data = meta_sub)
    mat_mm <- GetAssayData(so_sub, slot = 'counts') %*% mm
    keep_genes <- rowSums(mat_mm > 0) > 0
    mat_mm <- mat_mm[keep_genes, , drop = FALSE] %>% as.data.frame()
    
    ## Reset column names.
    ## The column names will be in the form "subjectID:group".
    colnames(mat_mm) <- gsub("replicate|label", "", colnames(mat_mm))
    
    ## Drop any empty columns
    keep_samples <- colSums(mat_mm) > 0
    mat_mm <- mat_mm[, keep_samples, drop = FALSE]
    return(mat_mm)
  }) %>% setNames(keep)
  
  ## Remove cell types that did not yield any pseudo-bulk matrix
  pseudobulks <- pseudobulks[!purrr::map_lgl(pseudobulks, is.null)]
  pseudobulks <- pseudobulks[!is.na(pseudobulks)]
  
  ## Remove cell types with too few genes
  min_dim <- purrr::map_dbl(pseudobulks, ~ nrow(as.data.frame(.)))
  pseudobulks <- pseudobulks[min_dim > 1]
  
  ## Also filter out cell types without replicates
  min_repl <- purrr::map_int(pseudobulks, function(x) {
    tmp <- as.data.frame(x)
    targets <- data.frame(group_sample = colnames(tmp)) %>%
      dplyr::mutate(group = gsub(".*\\:", "", group_sample))
    if (n_distinct(targets$group) == 1)
      return(0L)
    min(table(targets$group))
  })
  pseudobulks <- pseudobulks[min_repl >= 2]
  
  ## List to store DESeq2 objects
  DESeq2_objs <- list()
  
  ## Load the DESeq2 library
  library(DESeq2)
  
  ## Run DESeq2 for each cell type
  DE <- purrr::map2(pseudobulks, names(pseudobulks), function(x, cell_type) {
    tryCatch({
      ## Create a targets data.frame from the pseudo-bulk column names.
      ## We extract both the group and replicate information.
      targets <- data.frame(group_sample = colnames(x)) %>%
        dplyr::mutate(group = gsub(".*\\:", "", group_sample),
                      replicate = gsub("\\:.*", "", group_sample))
      
      ## If a covariate is provided, retrieve its value for each replicate.
      if (!is.null(covariate)) {
        ## Subset the original object to the current cell type.
        so_sub <- subset(seurat_object, idents = cell_type)
        covariate_df <- so_sub@meta.data %>%
          dplyr::select(replicate, covariate) %>%
          dplyr::distinct()
        targets <- targets %>% dplyr::left_join(covariate_df, by = "replicate")
      }
      
      ## If the original metadata label was a factor, preserve its levels here.
      if (is.factor(meta$label)) {
        targets$group <- factor(targets$group, levels = levels(meta$label))
      }
      
      ## Build the design formula.
      ## If a covariate is provided, include it in the design.
      if (!is.null(covariate)) {
        design_formula <- as.formula("~ covariate + group")
      } else {
        design_formula <- as.formula("~ group")
      }
      
      ## Build the DESeq2 dataset
      dds <- DESeqDataSetFromMatrix(countData = x,
                                    colData = targets,
                                    design = design_formula)
      
      ## Set default parameters if not provided
      if (!'test' %in% names(params))
        params$test <- 'LRT'
      if (!'fitType' %in% names(params))
        params$fitType <- 'parametric'
      if (!'sfType' %in% names(params))
        params$sfType <- 'poscounts'
      if (!'independentFiltering' %in% names(params))
        params$independentFiltering <- FALSE
      if (!'betaPrior' %in% names(params))
        params$betaPrior <- FALSE
      
      ## Run DESeq2 with either the Wald or LRT test
      if (params$test == 'Wald') {
        dds <- try(DESeq(dds,
                         test = params$test,
                         fitType = params$fitType,
                         sfType = params$sfType,
                         betaPrior = params$betaPrior))
      } else {
        dds <- try(DESeq(dds,
                         test = params$test,
                         reduced = ~ 1,
                         fitType = params$fitType,
                         sfType = params$sfType,
                         betaPrior = params$betaPrior))
      }
      
      res <- results(dds, independentFiltering = params$independentFiltering)
      result <- as.data.frame(res) %>%
        tibble::rownames_to_column('gene') %>%
        dplyr::mutate(test = 'pseudobulk_DESeq2')
      
      ## Save the DESeq2 object for later inspection if desired
      DESeq2_objs[[cell_type]] <<- dds
      
      return(result)
    }, error = function(e) {
      message("Error in cell type ", cell_type, ": ", e)
      return(data.frame())
    })
  })
  
  ## Standardize the column names to match Seurat output conventions using conditional renaming
  DE <- purrr::map(DE, function(df) {
    rename_map <- c('p.value' = 'p_val',
                    'padj' = 'p_val_adj',
                    'log2FoldChange' = 'avg_logFC')
    for (old_name in names(rename_map)) {
      if (old_name %in% colnames(df)) {
        colnames(df)[colnames(df) == old_name] <- rename_map[old_name]
      }
    }
    return(df)
  })
  
  return(list(DE = DE, pseudobulks = pseudobulks, DESeq2_objs = DESeq2_objs))
}
```

### Customize pathway labels

```{r}
customize_pathway_labels <- function(GSEAres, replace_underscores = FALSE) {
  # Ensure the Description column is a character vector
  GSEAres@result$Description <- as.character(GSEAres@result$Description)
  
  # Remove the "REACTOME_" prefix from the pathway IDs
  GSEAres@result$Description <- gsub("^REACTOME_", "", GSEAres@result$Description)
  
  # Convert the pathway descriptions to lowercase
  GSEAres@result$Description <- tolower(GSEAres@result$Description)
  
  # Function to replace terms based on a provided replacements list
  replace_terms <- function(text, replacements) {
    for (term in names(replacements)) {
      text <- gsub(term, replacements[term], text)
    }
    return(text)
  }
  
  # set labels
  labels <- c("fcgamma_receptor" = "", "fcgr" = "FCGR", "gm_csf" = "GM-CSF", "tcr" = "TCR", 
                      "ddx58" = "RIG-I", "ifih1" = "MDA-5", "alpha_beta" = "alpha/beta",
                      "interleukin_1" = "IL-1", "interleukin_3" = "IL-3", "interleukin_4" = "IL-4", 
                      "interleukin_5" = "IL-5", "interleukin_13" = "IL-13",
                      "mhc" = "MHC", "class_ii" = "class_II", "class_i_MHC" = "MHC_class_I", 
                      "jak_stat" = "JAK-STAT", "pd_1" = "PD-1", "tnf" = "TNF", 
                      "tgf" = "TGF", "TGF_beta" = "TGFb", "bcr" = "BCR", "cd28" = "CD28", "hiv" = "HIV", "b_cell" = "B_cell",
                      "sars_cov" = "SARS-CoV", "SARS-CoV_1" = "SARS-CoV-1", "SARS-CoV_2" = "SARS-CoV-2",
                      "hcmv" = "hCMV")
  
  # Apply the replacement function using the provided 'labels' vector
  GSEAres@result$Description <- replace_terms(GSEAres@result$Description, labels)
  
  # After applying replacements, optionally remove underscores by replacing them with spaces
  if (replace_underscores) {
    GSEAres@result$Description <- gsub("_", " ", GSEAres@result$Description)
  }
  
  # Capitalize the first letter of each pathway description
  GSEAres@result$Description <- paste(
    toupper(substring(GSEAres@result$Description, 1, 1)),
    substring(GSEAres@result$Description, 2),
    sep = ""
  )
  
  return(GSEAres)
}
```

### GSEA

```{r}
run_GSEA_analysis <- function(DEG_df,
                              collapse_pathways = FALSE,
                              pvalueCutoff    = 0.05,
                              gmt_path,
                              subset_pathways = NULL,
                              scoreType_GSEA = "std") {
  
  library(clusterProfiler)
  library(fgsea)
  library(dplyr)

  # 1) Read the GMT and subset to genes in your data
  pwl2 <- read.gmt(gmt_path)  
  bg_genes_REACTOME <- pwl2 %>% dplyr::filter(gene %in% DEG_df$gene)
  
  # 2) If subset_pathways is provided, filter to just those terms
  if (!is.null(subset_pathways)) {
    bg_genes_REACTOME <- bg_genes_REACTOME %>% dplyr::filter(term %in% subset_pathways)
  }
  
  # 3) Create a named, ranked vector of statistics
  gene_rankings_REACTOME <- DEG_df$stat
  names(gene_rankings_REACTOME) <- DEG_df$gene
  gene_rankings_REACTOME <- sort(na.omit(gene_rankings_REACTOME), decreasing = TRUE)

  if (!collapse_pathways) {
    # — Non-collapsed GSEA —
    GSEAres <- GSEA(gene_rankings_REACTOME,
                    TERM2GENE    = bg_genes_REACTOME,
                    minGSSize    = 10,
                    maxGSSize    = length(gene_rankings_REACTOME) - 1,
                    pvalueCutoff = pvalueCutoff,
                    pAdjustMethod= "BH",
                    by           = "fgsea",
                    scoreType    = scoreType_GSEA,
                    verbose      = FALSE,
                    eps          = 0)
    
  } else {
    # — Collapsed-pathway GSEA —
    
    # a) Load all pathways as a list
    gene_sets <- GSEABase::getGmt(gmt_path)
    # b) If subsetting, drop any you don’t want
    if (!is.null(subset_pathways)) {
      gene_sets <- gene_sets[ names(gene_sets) %in% subset_pathways ]
    }
    # c) Convert to a simple list of character vectors
    bg_genes_REACTOME_list <- lapply(gene_sets, GSEABase::geneIds)
    names(bg_genes_REACTOME_list) <- names(gene_sets)
    
    # d) Run fgsea (gives you the stats table)
    GSEAres <- fgsea(pathways   = bg_genes_REACTOME_list,
                     stats      = gene_rankings_REACTOME,
                     scoreType  = scoreType_GSEA,
                     minSize    = 10,
                     maxSize    = length(gene_rankings_REACTOME) - 1) %>%
      dplyr::filter(padj < pvalueCutoff)
    
    # e) Collapse redundant pathways
    cp <- collapsePathways(GSEAres, bg_genes_REACTOME_list, gene_rankings_REACTOME)
    
    # f) Keep only the “main” pathways
    GSEAres <- GSEAres %>% dplyr::filter(pathway %in% cp$mainPathways)
  }
  
  return(GSEAres)
}
```

### GSEA heatmap

```{r}
gsea_heatmap <- function(
  gsea_list,
  pval_cutoff = 0.05,
  collapsed_pathways = FALSE,
  pathways_show = NULL,
  title = "Pseudobulk GSEA Analysis",
  show_rownames = TRUE,
  rows_annotations = NULL,
  col_annotation_rows = NULL,
  col_annotation_cols = NULL,
  title_annotation_rows = NULL,
  title_annotation_cols = NULL,
  gap_by_row = FALSE,
  gap_by_col = FALSE
) {

  # Helper to extract & rename NES and padj/qvalue
  extract_gsea_data <- function(gsea_obj, group_name, collapsed_pathways) {
    id_col   <- ifelse(collapsed_pathways, "pathway", "ID")
    padj_col <- ifelse(collapsed_pathways, "padj",    "qvalue")

    gsea_obj %>%
      as.data.frame() %>%
      dplyr::select(all_of(id_col), NES, all_of(padj_col)) %>%
      dplyr::rename(
        !!paste0("NES_",   group_name) := NES,
        !!paste0("padj_", group_name) := .data[[padj_col]]
      )
  }

  # 1) build a merged data.frame of NES + padj
  gsea_data_list <- lapply(names(gsea_list), function(name) {
    extract_gsea_data(gsea_list[[name]], name, collapsed_pathways)
  })
  identifier <- if (collapsed_pathways) "pathway" else "ID"
  df <- Reduce(function(x, y) full_join(x, y, by = identifier), gsea_data_list)

  # 2) fill NAs in NES → 0
  nes_cols <- grep("^NES_", colnames(df), value = TRUE)
  df <- df %>% dplyr::mutate(across(all_of(nes_cols), ~ ifelse(is.na(.), 0, .)))

  # 3) fill NAs in padj → 1 & filter rows with any padj < cutoff
  padj_cols <- grep("^padj_", colnames(df), value = TRUE)
  df <- df %>%
    dplyr::mutate(across(all_of(padj_cols), ~ ifelse(is.na(.), 1, .))) %>%
    dplyr::filter(rowSums(dplyr::across(all_of(padj_cols), ~ . < pval_cutoff)) > 0)

  # 3b) subset to only the user-requested pathways (if any)
  if (!is.null(pathways_show)) {
    keep <- pathways_show
    matched <- df[[identifier]] %in% keep
    if (!any(matched)) {
      warning("No pathways in 'pathways_show' matched your data; plotting all filtered pathways instead.")
    } else {
      df <- df[matched, , drop = FALSE]
    }
  }
  
  # 4) pretty-print the pathway labels
  customize_labels <- function(labels) {
    labels <- as.character(labels)
    labels <- gsub("^REACTOME_", "", labels)
    labels <- tolower(labels)
    replacements <- c(
      "fcgamma_receptor" = "", "fcgr" = "FCGR", "gm_csf" = "GM-CSF", "tcr" = "TCR",
      "ddx58" = "RIG-I", "ifih1" = "MDA-5", "alpha_beta" = "alpha/beta",
      "interleukin_1" = "IL-1", "interleukin_3" = "IL-3", "interleukin_4" = "IL-4",
      "interleukin_5" = "IL-5", "interleukin_13" = "IL-13",
      "mhc" = "MHC", "class_ii" = "class_II", "class_i_MHC" = "MHC_class_I",
      "jak_stat" = "JAK-STAT", "pd_1" = "PD-1", "tnf" = "TNF",
      "tgf" = "TGF", "TGF_beta" = "TGFb", "bcr" = "BCR", "cd28" = "CD28",
      "hiv" = "HIV", "b_cell" = "B_cell",
      "sars_cov" = "SARS-CoV", "SARS-CoV_1" = "SARS-CoV-1", "SARS-CoV_2" = "SARS-CoV-2",
      "hcmv" = "hCMV", "csf" = "CSF"
    )
    for (term in names(replacements)) {
      labels <- gsub(term, replacements[term], labels)
    }
    labels <- gsub("_", " ", labels)
    labels <- paste0(toupper(substring(labels, 1, 1)), substring(labels, 2))
    labels
  }
  df[[identifier]] <- customize_labels(df[[identifier]])

  # 5) build NES-only matrix for heatmap
  df_NES <- df %>% dplyr::select(-all_of(padj_cols))
  rownames(df_NES) <- df_NES[[identifier]]
  df_NES[[identifier]] <- NULL
  df_NES <- as.matrix(df_NES)

  # 6) prep & reorder row-annotations if provided
  if (!is.null(rows_annotations)) {
    rownames(rows_annotations) <- customize_labels(rownames(rows_annotations))
    common_pathways <- intersect(rownames(df_NES), rownames(rows_annotations))
    if (length(common_pathways) == 0) {
      warning("No matching pathways found between heatmap data and rows_annotations.")
      rows_annotations <- NULL
    } else {
      rows_annotations <- rows_annotations[common_pathways, , drop = FALSE]
      group_var <- if ("Group" %in% colnames(rows_annotations)) rows_annotations$Group else rows_annotations[[1]]
      ord <- order(as.factor(group_var))
      df_NES <- df_NES[common_pathways[ord], , drop = FALSE]
      rows_annotations <- rows_annotations[common_pathways[ord], , drop = FALSE]
    }
  }

  # 7) format p-values for display
  p_values <- df %>% dplyr::select(all_of(padj_cols))
  rownames(p_values) <- df[[identifier]]
  formatted_p_values <- p_values %>%
    dplyr::mutate(across(everything(), ~ dplyr::case_when(
      . < 1e-4 ~ "****",
      . < 1e-3 ~ "***",
      . < 1e-2 ~ "**",
      . < 0.05 ~ "*",
      . < 0.1  ~ "(*)",
      TRUE     ~ ""
    )))
  formatted_p_values <- formatted_p_values[rownames(df_NES), , drop = FALSE]

  # 8) column-metadata & titles
  meta.heatmap <- data.frame(Meta1 = names(gsea_list), row.names = colnames(df_NES))
  if (!is.null(rows_annotations) && !is.null(title_annotation_rows)) {
    colnames(rows_annotations) <- title_annotation_rows
  }
  if (!is.null(title_annotation_cols)) {
    colnames(meta.heatmap) <- title_annotation_cols
  }

  # 9) build annotation_colors
  annotation_colors <- list()
  if (!is.null(rows_annotations) && !is.null(col_annotation_rows)) {
    if (is.list(col_annotation_rows)) {
      for (cn in colnames(rows_annotations)) {
        if (cn %in% names(col_annotation_rows)) annotation_colors[[cn]] <- col_annotation_rows[[cn]]
      }
    } else {
      annotation_colors[[colnames(rows_annotations)[1]]] <- col_annotation_rows
    }
  }
  if (!is.null(col_annotation_cols)) {
    if (is.list(col_annotation_cols)) {
      for (cn in colnames(meta.heatmap)) {
        if (cn %in% names(col_annotation_cols)) annotation_colors[[cn]] <- col_annotation_cols[[cn]]
      }
    } else {
      annotation_colors[[colnames(meta.heatmap)[1]]] <- col_annotation_cols
    }
  }

  # 10) compute gaps based on flags
  if (gap_by_row && !is.null(rows_annotations)) {
    grp_row <- if ("Group" %in% colnames(rows_annotations)) "Group" else colnames(rows_annotations)[1]
    f_row    <- factor(rows_annotations[[grp_row]], levels = unique(rows_annotations[[grp_row]]))
    gaps_row <- which(diff(as.numeric(f_row)) != 0)
  } else {
    gaps_row <- NULL
  }
  if (gap_by_col) {
    grp_col  <- colnames(meta.heatmap)[1]
    f_col    <- factor(meta.heatmap[[grp_col]], levels = unique(meta.heatmap[[grp_col]]))
    gaps_col <- which(diff(as.numeric(f_col)) != 0)
  } else {
    gaps_col <- NULL
  }

  # 11) draw the heatmap
  pheatmap::pheatmap(
    df_NES,
    clustering_method    = "complete",
    annotation_col       = meta.heatmap,
    annotation_row       = rows_annotations,
    annotation_colors    = if (length(annotation_colors) > 0) annotation_colors else NULL,
    annotation_names_col = FALSE,
    annotation_names_row = FALSE,
    show_rownames        = show_rownames,
    show_colnames        = FALSE,
    cluster_rows         = ifelse(is.null(rows_annotations), TRUE, FALSE),
    cluster_cols         = FALSE,
    display_numbers      = formatted_p_values,
    fontsize_number      = 10,
    fontsize_row         = 10,
    fontsize_col         = 10,
    angle_col            = "315",
    border_color         = NA,
    color                = colorRampPalette(c("#2171b5", "white", "#ef3b2c"))(100),
    breaks               = seq(-max(abs(df_NES)), max(abs(df_NES)), length.out = 101),
    gaps_row             = gaps_row,
    gaps_col             = gaps_col,
    heatmap_legend_param = list(title = "NES", title_position = "topcenter", legend_direction = "vertical"),
    main                 = title
  )
}
```

## Gene scores

```{r}
plot_gene_set_scores <- function(seurat_obj, cell_subset, annotation_column, gene_lists_focus,
                                 gene_lists_bg, bg_origin, pathway_titles = NULL, 
                                 tissue = c("CSF", "PBMC"), nrow = 1, calc_signif = TRUE,
                                 title_wrap_width = 40) {
  # If pathway_titles is not provided, use the names from gene_lists_focus
  if (is.null(pathway_titles)) {
    pathway_titles <- setNames(as.list(names(gene_lists_focus)), names(gene_lists_focus))
  }
  
  # Generate dynamic comparisons based on tissues provided.
  comparisons_list <- list()
  if (length(tissue) >= 2) {
    for (i in 1:(length(tissue) - 1)) {
      for (j in (i + 1):length(tissue)) {
        comparisons_list <- c(comparisons_list, list(c(paste0("control_", tissue[i]), 
                                                       paste0("control_", tissue[j]))))
      }
    }
    for (i in 1:(length(tissue) - 1)) {
      for (j in (i + 1):length(tissue)) {
        comparisons_list <- c(comparisons_list, list(c(paste0("PWH_", tissue[i]), 
                                                       paste0("PWH_", tissue[j]))))
      }
    }
  }
  for (t in tissue) {
    comparisons_list <- c(comparisons_list, list(c(paste0("control_", t), paste0("PWH_", t))))
  }
  
  # List to store plots
  gene_set_score_plots <- list()
  
  for (gene_list in names(gene_lists_focus)) {
    # Remove NA values from the current gene list
    current_genes <- na.omit(gene_lists_focus[[gene_list]])
    
    # Intersect with features present in the Seurat object
    current_genes <- intersect(current_genes, rownames(seurat_obj))
    
    # Check if there are enough valid genes to calculate a module score;
    # silently skip if not.
    if (length(current_genes) < 2) {
      next
    }
    
    # Subset by the specified cell type using the annotation_column
    seurat_obj_score_subset <- subset(seurat_obj, subset = !!sym(annotation_column) == cell_subset)
    
    # Filter cells to only include the specified tissues.
    selected_cells <- which(seurat_obj_score_subset@meta.data$tissue %in% tissue)
    seurat_obj_score_subset <- subset(seurat_obj_score_subset, cells = selected_cells)
    
    # Calculate module score using the filtered gene list
    seurat_obj_score_subset <- AddModuleScore(seurat_obj_score_subset, 
                                              features = list(current_genes), 
                                              name = "pathway_score")
    
    # Summarize the module score by grouping over tissue, HIV_status, and subject_id
    df <- seurat_obj_score_subset@meta.data %>% 
      group_by(tissue, HIV_status, subject_id) %>% 
      summarize(mean = mean(pathway_score1), .groups = 'drop')
    df$group <- paste0(df$HIV_status, "_", df$tissue)
    
    # Reorder the group factor to the desired order dynamically.
    # For each tissue provided, the order is "control_<tissue>", then "PWH_<tissue>"
    desired_levels <- unlist(lapply(tissue, function(x) c(paste0("control_", x), paste0("PWH_", x))))
    df$group <- factor(df$group, levels = desired_levels)
    
    # --- Title Processing ---
    title_text <- pathway_titles[[gene_list]]
    title_text <- gsub("^REACTOME_", "", title_text)
    title_text <- gsub("_", " ", title_text)
    
    # Convert to lower case then to Title Case (first letter of each word capitalized)
    title_text <- stringr::str_to_title(tolower(title_text))
    
    # Define custom replacements (keys are in lower case with spaces)
    replacements <- c("fcgamma_receptor" = "", "fcgr" = "FCGR", "gm_csf" = "GM-CSF", "tcr" = "TCR", 
                      "ddx58" = "RIG-I", "ifih1" = "MDA-5", "alpha_beta" = "alpha/beta",
                      "interleukin_1" = "IL-1", "interleukin_3" = "IL-3", "interleukin_4" = "IL-4", 
                      "interleukin_5" = "IL-5", "interleukin_13" = "IL-13",
                      "mhc" = "MHC", "class_ii" = "class_II", "class_i_MHC" = "MHC_class_I", 
                      "jak_stat" = "JAK-STAT", "pd_1" = "PD-1", "tnf" = "TNF", 
                      "tgf" = "TGF", "TGF_beta" = "TGFb", "bcr" = "BCR", "cd28" = "CD28", "hiv" = "HIV", "b_cell" = "B_cell",
                      "sars_cov" = "SARS-CoV", "SARS-CoV_1" = "SARS-CoV-1", "SARS-CoV_2" = "SARS-CoV-2",
                      "hcmv" = "hCMV")
    
    for (k in names(replacements)) {
      key_title <- stringr::str_to_title(k)
      pattern <- paste0("\\b", key_title, "\\b")
      title_text <- gsub(pattern, replacements[k], title_text)
      title_text <- gsub("\\bIi\\b", "II", title_text)
    }
    
    # Wrap the title using the custom width
    title_text <- stringr::str_wrap(title_text, width = title_wrap_width)
    # --- End Title Processing ---
    
    # Create plot for the current gene list
    p <- ggplot(df, aes(y = mean, x = group, fill = group)) +
      geom_boxplot(position = position_dodge(width = 1)) +
      geom_jitter(position = position_dodge(width = 1), alpha = 0.3)
    
    if (calc_signif) {
      p <- p + geom_signif(comparisons = comparisons_list, 
                           step_increase = 0.12, tip_length = 0, test = "wilcox.test", map_signif_level = TRUE)
    }
    
    p <- p + labs(x = "", y = "Gene Module Score", 
                  title = title_text, 
                  subtitle = paste0("Set of ", length(current_genes), "/", length(gene_lists_bg[[gene_list]]), " genes")) +
      theme_bw() +
      theme(plot.title = element_text(size = 12, hjust = 0.5, face = "bold"),
            plot.subtitle = element_text(size = 11, hjust = 0.5), 
            axis.text.x = element_blank(),
            axis.ticks.x = element_blank())
    
    gene_set_score_plots[[gene_list]] <- p
  }
  
  # Final plot annotation now only includes the cell subset information.
  final_plot <- wrap_plots(gene_set_score_plots, nrow = nrow, guides = "collect", axis_titles = "collect") +
    plot_annotation(paste0("Gene Scores in ", cell_subset))
  
  return(final_plot)
}
```
