## GEX analysis of all cells

### Load IG/TR genes to be removed from variable features

```{r}
# read in IG/TR features pulled from Ensembl 103 using biomaRt
load(file.path("QC_features_meta_human.RData"))

# create objects with genes to exclude
biotypes_excl <- unique(features_meta[["gene_biotype"]])[grepl(pattern = "^IG_|^TR_",
                                                               x = unique(features_meta[["gene_biotype"]]))]
remove_genes <- features_meta[["external_gene_name"]][features_meta[["gene_biotype"]] %in% biotypes_excl]
```

### Run Seurat pipeline without integration

```{r}
# run normalization and scaling
gex_choroid <- NormalizeData(gex_choroid, normalization.method = "LogNormalize", scale.factor = 10000)
gex_choroid <- FindVariableFeatures(gex_choroid, selection.method = "vst", nfeatures = 2000)
gex_choroid <- ScaleData(gex_choroid)

# remove Ig and Tr genes before clustering
bool_remove_genes <- VariableFeatures(gex_choroid) %in% remove_genes
VariableFeatures(gex_choroid) <- VariableFeatures(gex_choroid)[!bool_remove_genes]
cat("After removing IG/TR genes, the total number of variable features is:", length(VariableFeatures(gex_choroid)), "\n")

# dimensionality reduction
gex_choroid <- RunPCA(gex_choroid, npcs = 70)
gex_choroid <- RunUMAP(gex_choroid, dims = 1:50, reduction = "pca", reduction.name = "umap.unintegrated")

# clustering
gex_choroid <- FindNeighbors(gex_choroid, dims = 1:50, reduction = "pca")
gex_choroid <- FindClusters(gex_choroid, resolution = 0.5)

# check for batch effects
DimPlot(gex_choroid, group.by = "sample_name", reduction = "umap.unintegrated")
```

### Doublet identification (scDblFinder)

```{r}
# convert Seurat object to SingleCellExperiment
sce <- as.SingleCellExperiment(gex_choroid)

# run scDblFinder with different doublet rates
sce_def <- scDblFinder(sce, samples = "sample_name")                

# add info to seurat object
meta_scdblfinder_def <- sce_def@colData@listData %>% as.data.frame() %>% dplyr::select(scDblFinder.class)
rownames(meta_scdblfinder_def) <- sce_def@colData@rownames
gex_choroid <- AddMetaData(object = gex_choroid, metadata = meta_scdblfinder_def, col.name = "doublet_def")

# summarize doublet rates
gex_choroid@meta.data %>% 
  group_by(sample_name, doublet_def) %>% 
  summarize(count = n(), .groups = "drop") %>% 
  dplyr::filter(doublet_def == "doublet") %>%
  left_join(dbr_expected, by = "sample_name") %>%
  mutate(obs_dbr = count/total_count * 100,
         dbr_exp_obs_delta_0.08 = exp_dbr_0.08 - obs_dbr,
         dbr_exp_obs_delta_0.1 = exp_dbr_0.1 - obs_dbr) %>%
  kable() %>% kable_styling("striped")

# remove unnecessary objects
rm(sce, sce_def, meta_scdblfinder_def)
```

```{r}
# visualize doublets before removal
DimPlot(gex_choroid, reduction = "umap.unintegrated", raster = TRUE, label = FALSE, 
         cells.highlight = gex_choroid$cell_id_unique[gex_choroid$doublet_def == "doublet"]) + 
  scale_color_manual(values = c("Unselected" = "grey", "Group_1" = "red"),
                     labels = c("Unselected" = "Singlets", "Group_1" = "Doublets")) +
  ggtitle("Doublets") + NoLegend() +
  theme(plot.title = element_text(hjust = 0.5), axis.title = element_text(hjust = 0.5))

# plot proliferation markers
VlnPlot(gex_choroid, ncol = 3,
        features = c("TOP2A", "MKI67", "CDK1", "MS4A1", "CD79A", "IGKC", "FKBP11", "CD19"))
```

```{r}
# remove doublets for non B cells; B cell doublets will be removed more careful after subclustering
barcodes_remove <- gex_choroid@meta.data %>% dplyr::filter(doublet_def == "doublet" & seurat_clusters != 20) %>% pull(cell_id_unique)
paste0("Among ", length(Cells(gex_choroid)), " cells (after QC filtering), ", length(barcodes_remove), " (", round(length(barcodes_remove) / length(Cells(gex_choroid)) * 100, 1), "%) cells will be removed as doublets.")

# remove doublets
gex_choroid <- subset(gex_choroid, subset = cell_id_unique %in% barcodes_remove, invert = TRUE)

# remove objects
rm(barcodes_remove)
```

### Integration using Harmony

```{r}
# split layers
gex_choroid[["RNA"]] <- split(gex_choroid[["RNA"]], f = gex_choroid$sample_name)

# reprocess seurat object by sample
gex_choroid <- NormalizeData(gex_choroid, normalization.method = "LogNormalize", scale.factor = 10000, verbose = FALSE)
gex_choroid <- FindVariableFeatures(gex_choroid, selection.method = "vst", nfeatures = 2000, verbose = FALSE)
gex_choroid <- ScaleData(gex_choroid, verbose = FALSE)

# remove Ig and Tr genes before clustering
bool_remove_genes <- VariableFeatures(gex_choroid) %in% remove_genes
VariableFeatures(gex_choroid) <- VariableFeatures(gex_choroid)[!bool_remove_genes]
cat("After removing IG/TR genes, the total number of variable features is:", length(VariableFeatures(gex_choroid)), "\n")

# dimensionality reduction
gex_choroid <- RunPCA(gex_choroid, npcs = 70)

# perform integration
gex_choroid <- IntegrateLayers(
  object = gex_choroid, method = HarmonyIntegration,
  orig.reduction = "pca", new.reduction = "harmony",
  verbose = FALSE
)

# join layers without integration
gex_choroid[["RNA"]] <- JoinLayers(gex_choroid[["RNA"]])

# Run UMAP and clustering using harmony embeddings
gex_choroid <- RunUMAP(gex_choroid, dims = 1:50, reduction = "harmony", reduction.name = "umap.harmony")
gex_choroid <- FindNeighbors(gex_choroid, dims = 1:50, reduction = "harmony")
gex_choroid <- FindClusters(gex_choroid, resolution = seq(0.1, 1.6, 0.1))

# check for batch effects
p1 <- DimPlot(gex_choroid, group.by = "sample_name", reduction = "umap.unintegrated", raster = FALSE) + 
  labs(title = "No integration") + theme(plot.title = element_text(hjust = 0.5))
p2 <- DimPlot(gex_choroid, group.by = "sample_name", reduction = "umap.harmony", raster = FALSE) + 
  labs(title = "Harmony integration") + theme(plot.title = element_text(hjust = 0.5))
p1 + p2 + plot_layout(guides = "collect")
```

### Clustree

```{r}
clustree(gex_choroid, prefix = 'RNA_snn_res.', node_text_size = 3) + 
  labs (title = "Clustering Tree of all cells in Choroid Plexus data") +
  scale_color_manual(values = col_clustree) + 
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# set resolution
Idents(gex_choroid) <- gex_choroid$RNA_snn_res.0.3
gex_choroid$seurat_clusters <- gex_choroid$RNA_snn_res.0.3
```

### PCA

```{r}
# calculate % of explained variance
varExplained <- gex_choroid@reductions$pca@stdev^2 / gex_choroid@reductions$pca@misc$total.variance
pc1_percentage <- round(varExplained[1] * 100, 1)
pc2_percentage <- round(varExplained[2] * 100, 1)

# plot PCA
PCAPlot(gex_choroid, group.by = c("tissue", "sex", "age", "HIV_status"), raster = FALSE) & 
  xlab(paste0("\nPC_1 (", pc1_percentage, "%)")) & 
  ylab(paste0("PC_2 (", pc2_percentage,"%)\n"))

# remove unnecessary objects
rm(varExplained, pc1_percentage, pc2_percentage)
```

### UMAP

```{r}
# check for batch effects
p1 <- DimPlot(gex_choroid, reduction = "umap.harmony", group.by = "seurat_clusters", raster = FALSE, label = TRUE, label.box = FALSE) +
        scale_fill_manual(values = rep("#FFFFFF", length(gex_choroid$seurat_clusters))) +
        NoLegend()
p2 <- DimPlot(gex_choroid, reduction = "umap.harmony", group.by = "subject_id", raster = TRUE, label = FALSE)
p3 <- DimPlot(gex_choroid, reduction = "umap.harmony", group.by = "subject_id", split.by = "HIV_status", raster = TRUE, label = FALSE)
(p1 + p2) / p3 + plot_layout(guides = "collect")

# look at mixing based on HIV_status
DimPlot(gex_choroid, reduction = "umap.harmony", group.by = "HIV_status", raster = FALSE, label = FALSE)

# doublets
DimPlot(gex_choroid, reduction = "umap.harmony", group.by = "seurat_clusters", raster = FALSE, label = TRUE, 
         cells.highlight = gex_choroid$cell_id_unique[gex_choroid$doublet_def == "doublet"], sizes.highlight = 0.5) + 
  scale_color_manual(values = c("Unselected" = "grey", "Group_1" = "red"),
                     labels = c("Unselected" = "Singlets", "Group_1" = "Doublets")) +
  ggtitle("Doublets after integration", subtitle = "Only doublets in (unintegrated) B cell cluster kept") + 
  theme(axis.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))
```

### Cell type annotation

#### Load markers

```{r}
markers_all <- read.csv(header = TRUE, file = "markers.csv")
```

#### Plot cell markers

```{r}
p <- DotPlot(gex_choroid, cluster.idents = TRUE, group.by = "seurat_clusters",
        features = unique(markers_all$Marker)) %>%
  add_info_bar(method = "join", 
               info_type = "Cell_Type",
               info = markers_all %>%
                 dplyr::select(Cell_Type, Marker) %>%
                 dplyr::rename(features.plot = Marker),
               text_size = 6, label = FALSE) + RotatedAxis() +
  theme(strip.text.x = element_text(size = 12)) +
  ggtitle("All B cell markers")

# plot side metrics
plot_dot_side(p, gex_choroid, row_identity = "seurat_clusters",
              facet_col = "Cell_Type", info_to_add = c("cluster_size"))
```

#### Assign cluster annotations

```{r}
# annotations - resol 0.3
annotations_general <- rbind(c("0", "Oligodendrocytes_1"),        
                             c("1", "CP_mature_1"),           
                             c("2", "CP_mature_2"),             
                             c("3", "CP_mature_3"),           
                             c("4", "Microglia_1_APC"),     
                             c("5", "Astrocytes_1"),              
                             c("6", "CP_stroma"),                     
                             c("7", "T_CD8_CTL"),         
                             c("8", "Neuron_1_synaptic"),   
                             c("9", "Neuron_2_synaptic"),         
                             c("10", "CP_mature_4"),             
                             c("11", "OPC"),      
                             c("12", "CP_stroma_pericytes"),               
                             c("13", "B_cells"),        
                             c("14", "Endothelial"),             
                             c("15", "Astrocytes_2"),           
                             c("16", "Oligodendrocytes_2"),
                             c("17", "Erythrocytes"),        
                             c("18", "Neuron_2"),             
                             c("19", "Neuron_3"),           
                             c("20", "Microglia_2_non_APC"))           
colnames(annotations_general) <- c("Cluster", "CellType")
annotations_general <- data.frame(annotations_general)

# create object to use in RenameIdents
annotated_clusters <- setNames(annotations_general$CellType, annotations_general$Cluster)

# Rename Idents
Idents(gex_choroid) <- gex_choroid$seurat_clusters
gex_choroid <- RenameIdents(gex_choroid, annotated_clusters)
gex_choroid[["annotated_clusters"]] <- Idents(gex_choroid)
gex_choroid@meta.data$annotated_clusters <- factor(gex_choroid@meta.data$annotated_clusters, levels = sort(as.character(unique(gex_choroid@meta.data$annotated_clusters))))

# remove unnecessary objects
rm(annotations_general, annotated_clusters)
```

```{r}
# combine subcluster of the same cell type
gex_choroid@meta.data <- gex_choroid@meta.data %>%
                            mutate(annotated_clusters_comb = case_when(
                              annotated_clusters %in% c("Oligodendrocytes_1", "Oligodendrocytes_2") ~ "Oligodendrocytes",
                              annotated_clusters %in% c("CP_mature_1", "CP_mature_2", "CP_mature_3", "CP_mature_4") ~ "CP mature",
                              annotated_clusters %in% c("Microglia_1_APC", "Microglia_2_non_APC") ~ "Microglia",
                              annotated_clusters %in% c("Astrocytes_1", "Astrocytes_2") ~ "Astrocytes",
                              annotated_clusters %in% c("CP_stroma", "CP_stroma_pericytes") ~ "CP stroma",
                              annotated_clusters %in% c("Neuron_1_synaptic", "Neuron_2_synaptic", "Neuron_2", "Neuron_3") ~ "Neurons",
                              annotated_clusters %in% c("T_CD8_CTL") ~ "CD8 T cells",
                              annotated_clusters %in% c("B_cells") ~ "B cells",
                              TRUE ~ annotated_clusters))

# combine into broad cell types
gex_choroid@meta.data <- gex_choroid@meta.data %>%
                            mutate(annotated_clusters_gen = case_when(
                              annotated_clusters_comb %in% c("Oligodendrocytes", "OPC", "Astrocytes", "Neurons") ~ "Brain parenchymal cells",
                              annotated_clusters_comb %in% c("CP mature", "CP stroma") ~ "CP cells",
                              annotated_clusters_comb %in% c("CD8 T cells", "B cells", "Microglia") ~ "Immune cells",
                              TRUE ~ annotated_clusters_comb
                                ))

# set as final Idents
Idents(gex_choroid) <- gex_choroid$annotated_clusters_comb
```

##### Post-annotation quality check

```{r}
# plot UMAP with annotations
DimPlot(gex_choroid, reduction = "umap.harmony", group.by = "annotated_clusters", 
              raster = FALSE, label = TRUE, label.box = FALSE) + 
  scale_fill_manual(values = rep("#FFFFFF", length(gex_choroid$annotated_clusters))) +
  NoLegend()

# plot by sample_name
DimPlot(gex_choroid, reduction = "umap.harmony", group.by = "subject_id", raster = FALSE)

# plot by HIV status
DimPlot(gex_choroid, reduction = "umap.harmony", group.by = "annotated_clusters", split.by = "HIV_status", 
        raster = FALSE, label = FALSE, repel = TRUE) + NoLegend()

# cluster barplots for quality check
p1 <- plot_pcts(pcts = calc_pcts(data = gex_choroid[[]],
                           meta_group_by = c("annotated_clusters"),
                           focus_group = "subject_id"), 
          tissue_type = "Choroid Plexus", plot_type = "All", plot_value = "Subject_id",
          count_by = c("annotated_clusters"),
          x_axis = "annotated_clusters", x_axis_label = "Cell cluster", 
          fill_type = "subject_id", fill_label = "Subject ID", 
          details = "Resolution 0.3") +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(size = 12)) + 
  labels_rotate_x

p2 <- plot_pcts(pcts = calc_pcts(data = gex_choroid[[]],
                           meta_group_by = c("annotated_clusters"),
                           focus_group = "HIV_status"), 
          tissue_type = "", plot_type = "All", plot_value = "HIV Status",
          count_by = c("annotated_clusters"),
          x_axis = "annotated_clusters", x_axis_label = "Cell cluster", 
          fill_type = "HIV_status", fill_label = "HIV Status", 
          details = "Resolution 0.3") +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(size = 12)) + 
  labels_rotate_x


p1 / p2 + plot_layout(axes = "collect_x")
```

```{r}
# Dot plot of all cells
DotPlot(gex_choroid, cluster.idents = FALSE, group.by = "annotated_clusters",
        features = get_features_from_all(markers_df = markers_all, 
                                         cell_types = unique(markers_all$Cell_Type), alphabetize_all = FALSE)) %>%
  add_info_bar(method = "join", 
               info_type = "Cell_Type",
               info = markers_all %>%
                 dplyr::select(Cell_Type, Marker) %>%
                 dplyr::rename(features.plot = Marker),
               text_size = 6, label = FALSE) +
  theme(axis.text.x = element_text(size = 10, angle = 90)) +
  ggtitle("All Cell Type Markers") + NoLegend()

# Dot plot of myeloid cells
DotPlot(gex_choroid, cluster.idents = FALSE, group.by = "annotated_clusters",
        features = unique(markers_DC_microglia$Cell_Type)) %>%
  add_info_bar(method = "join", 
               info_type = "Cell_Type",
               info = markers_DC_microglia %>%
                 dplyr::select(Cell_Type, Marker) %>%
                 dplyr::rename(features.plot = Marker),
               text_size = 6, label = FALSE) + 
  labs(title = "Myeloid cells") + 
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 90)) + NoLegend()
```

### FindAllMarkers

```{r}
# FindAllMarkers
markers_clusters_all <- FindAllMarkers(gex_choroid, only.pos = TRUE)

# plot heatmap of top10 markers per cluster
markers_clusters_all %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10

DoHeatmap(subset(x = gex_choroid, downsample = 500), group.by = "annotated_clusters", slot = "scale.data", features = top10$gene) #+ NoLegend()

# remove object
rm(markers_seurat_B, top10)
```

### Remove viremic participants

```{r}
# remove viremic participants
gex_choroid_supppresed <- subset(gex_choroid, subset = HIV_status_det == "HIV_viremic", invert = TRUE)

# rerun UMAP for better visualization after filtering
gex_choroid_supppresed <- RunUMAP(gex_choroid_supppresed, reduction = "harmony", dims = 1:50)
```