## GEX analysis of all cells

### Load IG/TR genes to be removed from variable features

```{r}
# read in IG/TR features pulled from Ensembl 103 using biomaRt
load(file.path("QC_features_meta_human.RData"))

# create objects with genes to exclude
biotypes_excl <- unique(features_meta[["gene_biotype"]])[grepl(pattern = "^IG_|^TR_",
                                                               x = unique(features_meta[["gene_biotype"]]))]
remove_genes <- features_meta[["external_gene_name"]][features_meta[["gene_biotype"]] %in% biotypes_excl]
```

### Run Seurat pipeline with standard normalization

```{r}
# run normalization and scaling
gex_ART <- NormalizeData(gex_ART, normalization.method = "LogNormalize", scale.factor = 10000, verbose = FALSE)
gex_ART <- FindVariableFeatures(gex_ART, selection.method = "vst", nfeatures = 2000, verbose = FALSE) 
gex_ART <- ScaleData(gex_ART, verbose = FALSE)

# remove IG and TR genes before clustering
bool_remove_genes <- VariableFeatures(gex_ART) %in% remove_genes
VariableFeatures(gex_ART) <- VariableFeatures(gex_ART)[!bool_remove_genes]
cat("After removing IG/TR genes, the total number of variable features is:", length(VariableFeatures(gex_ART)), "\n")

# dimensionality reduction
gex_ART <- RunPCA(gex_ART, npcs = 70)
gex_ART <- RunUMAP(gex_ART, dims = 1:50)

# clustering
gex_ART <- FindNeighbors(gex_ART, dims = 1:50)
gex_ART <- FindClusters(gex_ART, resolution = 0.5)

# remove objects
rm(features_meta, biotypes_excl, bool_remove_genes, remove_genes)
```

### Doublet identification using scDblFinder

```{r}
# convert Seurat object to SingleCellExperiment
sce <- as.SingleCellExperiment(gex_ART)

# run scDblFinder
sce_def <- scDblFinder(sce, samples = "sample_name")

# add info to seurat object
meta_scdblfinder_def <- sce_def@colData@listData %>% as.data.frame() %>% dplyr::select(scDblFinder.class)
rownames(meta_scdblfinder_def) <- sce_def@colData@rownames
gex_ART <- AddMetaData(object = gex_ART, metadata = meta_scdblfinder_def, col.name = "doublet_def")

# calculate expected doublet rate based on cell counts before QC filtering
dbr_expected <- gex_ART_pre_QC@meta.data %>% 
                      group_by(sample_name) %>% 
                      summarize(total_count = n(), .groups = "drop") %>% 
                      mutate(exp_dbr_0.08 = total_count*0.8/1000)   # for a rate of 0.8% per 1000 captured cells (10X estimation)

# summarize doublet rates
gex_ART@meta.data %>% 
  group_by(sample_name, doublet_def) %>% 
  summarize(count = n(), .groups = "drop") %>% 
  dplyr::filter(doublet_def == "doublet") %>%
  left_join(dbr_expected, by = "sample_name") %>%
  mutate(obs_dbr = count/total_count * 100,
         dbr_exp_obs_delta_0.08 = exp_dbr_0.08 - obs_dbr,
         dbr_exp_obs_delta_0.1 = exp_dbr_0.1 - obs_dbr)

# remove unnecessary objects
rm(sce, sce_def, sce_0.006, meta_scdblfinder_def, gex_ART_pre_QC, dbr_expected)
```

### Doublet removal and reclustering

```{r}
# visualize doublets before removal
UMAPPlot(gex_ART, group.by = "seurat_clusters", raster = TRUE, label = TRUE, 
         cells.highlight = gex_ART$cell_id_unique[gex_ART$doublet_def == "doublet"]) + 
  scale_color_manual(values = c("Unselected" = "grey", "Group_1" = "red"),
                     labels = c("Unselected" = "Singlets", "Group_1" = "Doublets")) +
  labs(title = "Doublet Identification") + 
  theme(plot.title = element_text(hjust = 0.5), axis.title = element_text(hjust = 0.5))

# plot proliferation and plasmablast markers
# proliferating cells and plasmablasts can be erroneously classified as doublets
FeaturePlot(gex_ART, ncol = 4,
        features = c("TOP2A", "MKI67", "CDK1", "MS4A1", "CD79A", "IGKC", "FKBP11", "CD19"))
```

```{r}
# remove doublets for non B cells
# B cell doublets will be removed after sub-clustering to avoid removing plasmablasts that were misclassified as doublets
filter_barcodes <- gex_ART@meta.data %>% dplyr::filter(doublet_def == "doublet" & !seurat_clusters %in% c(5, 11, 28, 32)) %>% pull(cell_id_unique)

# remove doublets
gex_ART <- subset(gex_ART, subset = cell_id_unique %in% filter_barcodes, invert = TRUE)

# clustering
gex_ART <- FindNeighbors(gex_ART, dims = 1:50)
gex_ART <- FindClusters(gex_ART, resolution = seq(0.1, 1.6, 0.1))

# remove objects
rm(filter_barcodes)
```

### Clustree

```{r}
clustree(gex_ART, prefix = 'RNA_snn_res.', node_text_size=3) + 
  labs (title = "Clustering Tree of all cells in Blood and CSF") +
  scale_color_manual(values = col_clustree)
```

```{r}
# set resolution at 0.5
# resolution is chosen after iterative refinement based on the clustering tree and marker genes
Idents(gex_ART) <- gex_ART$RNA_snn_res.0.5
gex_ART$seurat_clusters <- gex_ART$RNA_snn_res.0.5
```

### PCA

```{r}
# calculate % of explained variance
varExplained <- gex_ART@reductions$pca@stdev^2 / gex_ART@reductions$pca@misc$total.variance
pc1_percentage <- round(varExplained[1] * 100, 1)
pc2_percentage <- round(varExplained[2] * 100, 1)

# plot PCA
PCAPlot(gex_ART, group.by = c("tissue", "sex", "age", "ethnicity", "race", "HIV_status"), raster = FALSE) & 
  xlab(paste0("\nPC_1 (", pc1_percentage, "%)")) & 
  ylab(paste0("PC_2 (", pc2_percentage,"%)\n"))

# remove unnecessary objects
rm(varExplained, pc1_percentage, pc2_percentage)
```

### UMAP

```{r UMAP}
# plot UMAP by cluster and by sample
p1 <- UMAPPlot(gex_ART, group.by = "seurat_clusters", raster = FALSE, label = TRUE, label.box = TRUE) + 
  scale_fill_manual(values = rep("#FFFFFF", length(gex_ART$seurat_clusters))) +
  NoLegend()
p2 <- UMAPPlot(gex_ART, group.by = "subject_id", raster = TRUE, label = FALSE)
p3 <- UMAPPlot(gex_ART, group.by = "subject_id", split.by = "HIV_status", raster = TRUE, label = FALSE)
(p1 + p2) / p3 + plot_layout(guides = "collect")

# look at mixing based on HIV_status
UMAPPlot(gex_ART, group.by = "HIV_status", raster = TRUE, label = FALSE)

# doublets
UMAPPlot(gex_ART, group.by = "seurat_clusters", raster = TRUE, label = TRUE, 
         cells.highlight = gex_ART$cell_id_unique[gex_ART$doublet_def == "doublet"]) + 
  scale_color_manual(values = c("Unselected" = "grey", "Group_1" = "red"),
                     labels = c("Unselected" = "Singlets", "Group_1" = "Doublets")) +
  ggtitle("Doublets default") + theme(axis.title = element_text(hjust = 0.5))

# remove objects
rm(p1, p2, p3, p4, p5)
```

### Cell type annotation

#### Automated annotation

```{r}
# run Azimuth-based annotation for cross-check of manual annotation later
gex_ART <- RunAzimuth(gex_ART, reference = "pbmcref")
```

#### Manual annotation

```{r}
# load marker genes
markers_all <- read.csv(header = TRUE, file = "markers_all.csv")
```

```{r}
p <- DotPlot(gex_ART, cluster.idents = TRUE, group.by = "seurat_clusters",
        features = unique(markers_all$Marker)) %>%
  add_info_bar(method = "join", 
               info_type = "Cell_Type",
               info = markers_all %>%
                 dplyr::select(Cell_Type, Marker) %>%
                 dplyr::rename(features.plot = Marker),
               text_size = 6, label = FALSE) + RotatedAxis() +
  ggtitle("All Cell Type Markers")

# plot side metrics
plot_dot_side(p, gex_ART, row_identity = "seurat_clusters",
              facet_col = "Cell_Type", info_to_add = c("cluster_size"))
```

```{r}
# annotations - resol 0.5
annotations_general <- rbind(c("0", "CD4_TCM_1"),   
                             c("1", "CD4_TCM_2"),  
                             c("2", "CD8_CTL_1"),    
                             c("3", "Mono_CD14_1"),     
                             c("4", "B_Naive"),   
                             c("5", "NK"),
                             c("6", "CD8_CTL_2"),  
                             c("7", "CD8_CTL_3"),  
                             c("8", "T_Naive"),   
                             c("9", "Mono_CD16"),   
                             c("10", "Treg"),  
                             c("11", "B_Memory"),   
                             c("12", "MAIT"),      
                             c("13", "CD4_TCM_3"),   
                             c("14", "Undef_Mono_T"),    
                             c("15", "cDC_CD1C"),   
                             c("16", "Microglia-like_1"),   
                             c("17", "Undef_T_CM"),    # check for low quality
                             c("18", "PLT"),    
                             c("19", "NK_CD56bright"),   
                             c("20", "T_PLT"),       # possible doublets not detected by doublet finder (PLT are transcriptionally quiescent)
                             c("21", "Mono_CD14_2"),       
                             c("22", "Microglia-like_2"),   
                             c("23", "pDC"),       
                             c("24", "Proliferating"),   
                             c("25", "Mono_CD14_4"),    # check for low quality
                             c("26", "Mono_CD14_3"),  
                             c("27", "Plasmablasts"),
                             c("28", "HPC"),
                             c("29", "cDC_CLEC9A"))
colnames(annotations_general) <- c("Cluster", "CellType")
annotations_general <- data.frame(annotations_general)

# create object to use in RenameIdents
annotated_clusters_all <- setNames(annotations_general$CellType, annotations_general$Cluster)

# Rename Idents
Idents(gex_ART) <- gex_ART$seurat_clusters
gex_ART <- RenameIdents(gex_ART, annotated_clusters_all)
gex_ART[["annotated_clusters_all"]] <- Idents(gex_ART)
gex_ART@meta.data$annotated_clusters_all <- factor(gex_ART@meta.data$annotated_clusters_all, levels = sort(as.character(unique(gex_ART@meta.data$annotated_clusters_all))))

# remove unnecessary objects
rm(annotations_general, annotated_clusters_all)
```

#### Check for low quality cells and misclustering

```{r}
# plot QC metrics for each cluster to screen for low quality cell clustering (undefined clusters)
VlnPlot(subset(gex_ART, subset = annotated_clusters_all %in% c("T_prolif", "Plasmablasts"), invert = TRUE), 
        features = c("percent.mt", "percent.ribo", "nCount_RNA", "nFeature_RNA"), 
        group.by = "annotated_clusters_all", ncol = 1, pt.size = 0) + plot_layout(axes = "collect_x", axis_titles = "collect_x")
## some clusters (including ambiguous labels) are of low quality and therefore will be removed from further analysis
```

```{r}
# check markers genes for automated and manual annotation
# automated annotation (Azimuth)
p <- DotPlot(gex_ART, cluster.idents = TRUE, group.by = "predicted.celltype.l2",
        features = unique(markers_all$Marker)) %>%
  add_info_bar(method = "join", 
               info_type = "Cell_Type",
               info = markers_all %>%
                 dplyr::select(Cell_Type, Marker) %>%
                 dplyr::rename(features.plot = Marker),
               text_size = 6, label = FALSE) + RotatedAxis() +
  ggtitle("All Cell Type Markers")

plot_dot_side(p, gex_ART, row_identity = "predicted.celltype.l2",
              facet_col = "Cell_Type", info_to_add = c("cluster_size"))

# manual annotation 
p <- DotPlot(gex_ART, cluster.idents = TRUE, group.by = "annotated_clusters_all",
        features = unique(markers_all$Marker)) %>%
  add_info_bar(method = "join", 
               info_type = "Cell_Type",
               info = markers_all %>%
                 dplyr::select(Cell_Type, Marker) %>%
                 dplyr::rename(features.plot = Marker),
               text_size = 6, label = FALSE) + RotatedAxis() +
  ggtitle("All Cell Type Markers")

plot_dot_side(p, gex_ART, row_identity = "annotated_clusters_all",
              facet_col = "Cell_Type", info_to_add = c("cluster_size"))

# check distribution of automated annotations among manual annotations
gex_ART@meta.data %>% dplyr::select(predicted.celltype.l2, annotated_clusters_all) %>% table()
```

#### Refine cell annotations

```{r}
# remove low quality clusters and doublet cluster
gex_ART <- subset(gex_ART, subset = annotated_clusters_all %in% c("Mono_CD14_4", "Undef_T_CM", "Undef_Mono_T", "T_PLT"), invert = TRUE)

# rerun UMAP after filtering
gex_ART <- RunUMAP(gex_ART, dims = 1:50)
```

```{r}
# merge clusters of the same cell type
# refined subtypes are out of the scope of this work and would necessitate subclustering
gex_ART@meta.data <- gex_ART@meta.data %>%
                          mutate(annotated_clusters_all_ref = case_when(
                                          annotated_clusters_all %in% c("CD4_TCM_1", "CD4_TCM_2", "CD4_TCM_3") ~ "CD4_TCM",
                                          annotated_clusters_all %in% c("CD8_CTL_1", "CD8_CTL_2", "CD8_CTL_3") ~ "CD8_CTL",
                                          annotated_clusters_all %in% c("B_Memory", "B_Naive") ~ "B",
                                          annotated_clusters_all %in% c("Mono_CD14_1", "Mono_CD14_2", "Mono_CD14_3") ~ "Mono_CD14",
                                          annotated_clusters_all %in% c("cDC_CD1C", "cDC_CLEC9A") ~ "cDC",
                                          annotated_clusters_all %in% c("Microglia-like_1", "Microglia-like_2") ~ "Microglia-like",
                                          TRUE ~ annotated_clusters_all))

# add high-level annotation layer for basic cell types
gex_ART@meta.data <- gex_ART@meta.data %>%
                          mutate(annotated_clusters_gen = case_when(
                                          annotated_clusters_all_ref %in% c("CD4_TCM", "T_Naive", "Treg", "MAIT", "CD8_CTL") ~ "T",
                                          annotated_clusters_all_ref %in% c("Mono_CD14", "Mono_CD16") ~ "Mono",
                                          annotated_clusters_all_ref %in% c("B_Memory", "B_Naive", "Plasmablasts") ~ "B",
                                          annotated_clusters_all_ref %in% c("cDC", "pDC") ~ "DC",
                                          annotated_clusters_all_ref %in% c("NK", "NK_CD56bright") ~ "NK",
                                          TRUE ~ annotated_clusters_all_ref))

# Rename cell annotations for better visualization
gex_ART$annotated_clusters_all_ref[gex_ART$annotated_clusters_all_ref == "CD4_TCM"] <- "CD4 TCM"
gex_ART$annotated_clusters_all_ref[gex_ART$annotated_clusters_all_ref == "CD8_CTL"] <- "CD8 CTL"
gex_ART$annotated_clusters_all_ref[gex_ART$annotated_clusters_all_ref == "B"] <- "B"
gex_ART$annotated_clusters_all_ref[gex_ART$annotated_clusters_all_ref == "Mono_CD14"] <- "CD14 Monocytes"
gex_ART$annotated_clusters_all_ref[gex_ART$annotated_clusters_all_ref == "Mono_CD16"] <- "CD16 Monocytes"
gex_ART$annotated_clusters_all_ref[gex_ART$annotated_clusters_all_ref == "T_Naive"] <- "Naive T"
gex_ART$annotated_clusters_all_ref[gex_ART$annotated_clusters_all_ref == "NK_CD56bright"] <- "NK CD56"
gex_ART$annotated_clusters_all_ref[gex_ART$annotated_clusters_all_ref == "PLT"] <- "Platelets"

# convert to factor
gex_ART$annotated_clusters_all_ref <- as.factor(gex_ART$annotated_clusters_all_ref)
gex_ART$annotated_clusters_gen <- as.factor(gex_ART$annotated_clusters_gen)

# set default Idents
Idents(gex_ART) <- gex_ART$annotated_clusters_all_ref
```

### B cell statistics

```{r}
# plot UMAP with highlighted BCRs
UMAPPlot(gex_ART, pt.size = 0.1, sizes.highlight = 0.1, raster = FALSE, repel = TRUE,
         group.by = "annotated_clusters_all_ref", label = TRUE, label.box = TRUE,
         cells.highlight = Cells(gex_ART)[which(gex_ART$contains_bcr)]) + 
    scale_fill_manual(values = rep("#FFFFFF", length(gex_ART$annotated_clusters_all_ref))) +
    NoLegend() + ggtitle("BCRs highlighted")

# calculate BCRs in B cell clusters
B_cells_df <- gex_ART@meta.data %>% dplyr::filter(annotated_clusters_all_ref %in% c("B", "Plasmablasts"))
BCR_neg_in_B_PBMC <- B_cells_df %>% dplyr::filter(contains_bcr == FALSE & tissue == "PBMC") %>% pull(cell_id_unique) %>% n_distinct()
BCR_neg_in_B_CSF <- B_cells_df %>% dplyr::filter(contains_bcr == FALSE & tissue == "CSF") %>% pull(cell_id_unique) %>% n_distinct()
nb_B_PBMC <- B_cells_df %>% dplyr::filter(tissue == "PBMC") %>% pull(cell_id_unique) %>% n_distinct()
nb_B_CSF <- B_cells_df %>% dplyr::filter(tissue == "CSF") %>% pull(cell_id_unique) %>% n_distinct()

# calculate proportions
paste0("Out of ", sum(gex_ART$contains_bcr), " cells with BCR, ", 
       sum(B_cells_df$contains_bcr), " (", round(sum(B_cells_df$contains_bcr) / sum(gex_ART$contains_bcr)*100,1), "%) are found in B cell clusters.")
paste0("Out of ", length(B_cells_df$contains_bcr), " cells in B cell clusters, ", 
       sum(B_cells_df$contains_bcr), " (", round(sum(B_cells_df$contains_bcr) / length(B_cells_df$contains_bcr)*100,1), "%) have BCRs.")
paste0("Out of ", nb_B_PBMC, " B cells in PBMCs, ", BCR_neg_in_B_PBMC, " (", round(BCR_neg_in_B_PBMC/nb_B_PBMC*100,1), "%) don't have BCRs.")
paste0("Out of ", nb_B_CSF, " B cells in CSF, ", BCR_neg_in_B_CSF, " (", round(BCR_neg_in_B_CSF/nb_B_CSF*100,1), "%) don't have BCRs.")

# remove objects
rm(B_cells_df, BCR_neg_in_B_PBMC, BCR_neg_in_B_CSF, nb_B_PBMC, nb_B_CSF)
```
