## GEX analysis of all cells

### Load IG/TR genes to be removed from variable features

```{r}
# read in IG/TR features pulled from Ensembl 103 using biomaRt
load(file.path("QC_features_meta_human.RData"))

# create objects with genes to exclude
biotypes_excl <- unique(features_meta[["gene_biotype"]])[grepl(pattern = "^IG_|^TR_",
                                                               x = unique(features_meta[["gene_biotype"]]))]
remove_genes <- features_meta[["external_gene_name"]][features_meta[["gene_biotype"]] %in% biotypes_excl]
```

### Run Seurat pipeline with standard normalization

```{r}
# run normalization and scaling
gex_multi <- NormalizeData(gex_multi, normalization.method = "LogNormalize", scale.factor = 10000)
gex_multi <- FindVariableFeatures(gex_multi, selection.method = "vst", nfeatures = 2000) 
gex_multi <- ScaleData(gex_multi)

# remove Ig and Tr genes before clustering
bool_remove_genes <- VariableFeatures(gex_multi) %in% remove_genes
VariableFeatures(gex_multi) <- VariableFeatures(gex_multi)[!bool_remove_genes]
cat("After removing IG/TR genes, the total number of variable features is:", length(VariableFeatures(gex_multi)), "\n")

# dimensionality reduction
gex_multi <- RunPCA(gex_multi, npcs = 70)
gex_multi <- RunUMAP(gex_multi, dims = 1:50, reduction = "pca")

# check for batch effects
DimPlot(gex_multi, group.by = "sample_name", reduction = "umap")
```

### Doublet identification (scDblFinder)

```{r}
# convert Seurat object to SingleCellExperiment
sce <- as.SingleCellExperiment(gex_multi)

# run scDblFinder
sce_def <- scDblFinder(sce, samples = "sample_name")

# add info to seurat object
meta_scdblfinder_def <- sce_def@colData@listData %>% as.data.frame() %>% dplyr::select(scDblFinder.class)
rownames(meta_scdblfinder_def) <- sce_def@colData@rownames
gex_multi <- AddMetaData(object = gex_multi, metadata = meta_scdblfinder_def, col.name = "doublet_def")

# calculate expected doublet rate
dbr_expected <- gex_multi_pre_QC@meta.data %>% 
                      group_by(sample_name) %>% 
                      summarize(total_count = n(), .groups = "drop") %>% 
                      mutate(exp_dbr_0.08 = total_count*0.8/1000,    # for a rate of 0.8% per 1000 captured cells
                             exp_dbr_0.1 = total_count*1/1000)       # for a rate of 1% per 1000 captured cells

# summarize doublet rates
gex_multi@meta.data %>% 
  group_by(sample_name, doublet_def) %>% 
  summarize(count = n(), .groups = "drop") %>% 
  dplyr::filter(doublet_def == "doublet") %>%
  left_join(dbr_expected, by = "sample_name") %>%
  mutate(obs_dbr = count/total_count * 100,
         dbr_exp_obs_delta_0.08 = exp_dbr_0.08 - obs_dbr,
         dbr_exp_obs_delta_0.1 = exp_dbr_0.1 - obs_dbr)

# remove unnecessary objects
rm(sce, sce_def, meta_scdblfinder_def, gex_multi_pre_QC, dbr_expected)
```

### Integration using Harmony

```{r}
# split layers
gex_multi[["RNA"]] <- split(gex_multi[["RNA"]], f = gex_multi$subject_id)

# reprocess seurat object by sample
gex_multi <- NormalizeData(gex_multi, normalization.method = "LogNormalize", scale.factor = 10000)
gex_multi <- FindVariableFeatures(gex_multi, selection.method = "vst", nfeatures = 2000)
gex_multi <- ScaleData(gex_multi)

# remove Ig and Tr genes before clustering
bool_remove_genes <- VariableFeatures(gex_multi) %in% remove_genes
VariableFeatures(gex_multi) <- VariableFeatures(gex_multi)[!bool_remove_genes]
cat("After removing IG/TR genes, the total number of variable features is:", length(VariableFeatures(gex_multi)), "\n")

# dimensionality reduction
gex_multi <- RunPCA(gex_multi, npcs = 70)

# perform integration
gex_multi <- IntegrateLayers(
  object = gex_multi, method = HarmonyIntegration,
  orig.reduction = "pca", new.reduction = "harmony",
  verbose = FALSE
)

# join layers without integration
gex_multi[["RNA"]] <- JoinLayers(gex_multi[["RNA"]])

# Run UMAP and clustering using harmony embeddings
gex_multi <- RunUMAP(gex_multi, dims = 1:50, reduction = "harmony", reduction.name = "umap.harmony")
gex_multi <- FindNeighbors(gex_multi, dims = 1:50, reduction = "harmony")
gex_multi <- FindClusters(gex_multi, resolution = 1.6)

# check for batch effects
p1 <- DimPlot(gex_multi, group.by = "sample_name", reduction = "umap", raster = TRUE) + 
  labs(title = "No integration") + theme(plot.title = element_text(hjust = 0.5))
p2 <- DimPlot(gex_multi, group.by = "sample_name", reduction = "umap.harmony", raster = TRUE) + 
  labs(title = "Harmony integration") + theme(plot.title = element_text(hjust = 0.5))
p1 + p2 + plot_layout(guides = "collect")
```

### Doublet removal & reclustering

```{r}
# visualize doublets before removal
DimPlot(gex_multi, reduction = "umap.harmony", group.by = "seurat_clusters", raster = TRUE, label = TRUE, 
         cells.highlight = gex_multi$cell_id_unique[gex_multi$doublet_def == "doublet"]) + 
  scale_color_manual(values = c("Unselected" = "grey", "Group_1" = "red"),
                     labels = c("Unselected" = "Singlets", "Group_1" = "Doublets")) +
  ggtitle("Doublet Identification") + theme(axis.title = element_text(hjust = 0.5))

# plot plasmablast markers
VlnPlot(gex_multi, features = c("MKI67", "IGKC", "FKBP11", "MZB1", "JCHAIN", "IRF4", "CD27", "SEC11C", "XBP1"), ncol = 1, pt.size = 0) 
# plasmablasts may be miclassified as doublets; therefore these cluster are excluded from doublet removal
```

```{r}
# remove doublets except for plasma cells, which can be misclassified as doublets by the software
remove_barcodes <- gex_multi@meta.data %>% dplyr::filter(doublet_def == "doublet" & !seurat_clusters %in% c(30)) %>% pull(cell_id_unique) 

# remove doublets
gex_multi <- subset(gex_multi, subset = cell_id_unique %in% remove_barcodes, invert = TRUE)

# Recluster using harmony embeddings before annotation
gex_multi <- RunUMAP(gex_multi, dims = 1:50, reduction = "harmony", reduction.name = "umap.harmony")
gex_multi <- FindNeighbors(gex_multi, dims = 1:50, reduction = "harmony")
gex_multi <- FindClusters(gex_multi, resolution = seq(0.1, 1.6, 0.1))

# remove objects
rm(remove_barcodes)
```

### Clustree

```{r}
clustree(gex_multi, prefix = 'RNA_snn_res.', node_text_size=3) + 
  labs (title = "Clustering Tree of all cells in CSF - Multiome") +
  scale_color_manual(values = col_clustree)
```

```{r}
# set resolution
Idents(gex_multi) <- gex_multi$RNA_snn_res.0.8
gex_multi$seurat_clusters <- gex_multi$RNA_snn_res.0.8
```

### PCA

```{r}
# calculate % of explained variance
varExplained <- gex_multi@reductions$pca@stdev^2 / gex_multi@reductions$pca@misc$total.variance
pc1_percentage <- round(varExplained[1] * 100, 1)
pc2_percentage <- round(varExplained[2] * 100, 1)

# plot PCA
PCAPlot(gex_multi, group.by = c("tissue", "sex", "age", "ethnicity", "race", "HIV_status"), raster = FALSE) & 
  xlab(paste0("\nPC_1 (", pc1_percentage, "%)")) & 
  ylab(paste0("PC_2 (", pc2_percentage,"%)\n"))

# remove unnecessary objects
rm(varExplained, pc1_percentage, pc2_percentage)
```

### UMAP

```{r}
# check for batch effects
p1 <- DimPlot(gex_multi, reduction = "umap.harmony", group.by = "seurat_clusters", raster = FALSE, label = TRUE, label.box = FALSE) +
        scale_fill_manual(values = rep("#FFFFFF", length(gex_multi$seurat_clusters))) +
        NoLegend()
p2 <- DimPlot(gex_multi, reduction = "umap.harmony", group.by = "subject_id", raster = TRUE, label = FALSE)
p3 <- DimPlot(gex_multi, reduction = "umap.harmony", group.by = "subject_id", split.by = "tissue", raster = TRUE, label = FALSE)
(p1 + p2) / p3 + plot_layout(guides = "collect")

# look at mixing based on HIV_status
DimPlot(gex_multi, reduction = "umap.harmony", group.by = "HIV_status", raster = FALSE, label = FALSE)

# visualizing remaining classified as doublets (proliferating and B cells)
DimPlot(gex_multi, reduction = "umap.harmony", group.by = "seurat_clusters", raster = TRUE, label = TRUE, 
         cells.highlight = gex_multi$cell_id_unique[gex_multi$doublet_def == "doublet"], sizes.highlight = 0.5) + 
  scale_color_manual(values = c("Unselected" = "grey", "Group_1" = "red"),
                     labels = c("Unselected" = "Singlets", "Group_1" = "Doublets")) +
  ggtitle("Doublets after integration", subtitle = "Only doublets in (unintegrated) B cell cluster kept") + 
  theme(axis.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))
```

### Cell type annotation

#### Find differential gene markers

### FindMarkers

```{r}
# FindAllMarkers
markers_clusters_all <- FindAllMarkers(gex_multi, only.pos = TRUE, group.by = "seurat_clusters")

# plot heatmap of top10 markers per cluster
markers_clusters_all %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    arrange(p_val_adj) %>%
    slice_head(n = 50) %>%
    ungroup() -> top10

DoHeatmap(subset(x = gex_multi, downsample = 500), group.by = "seurat_clusters", slot = "scale.data", features = top10$gene) + NoLegend()
```

### Automated annotation

```{r}
gex_multi <- RunAzimuth(gex_multi, reference = "pbmcref")
```

### Manual annotation

```{r}
# load marker genes
markers_all <- read.csv(header = TRUE, file = file.path(path_markers, "markers_all.csv"))
```

```{r}
p <- DotPlot(gex_multi, cluster.idents = TRUE, group.by = "seurat_clusters",
        features = unique(markers_all$Marker)) %>%
  add_info_bar(method = "join", 
               info_type = "Cell_Type",
               info = markers_all %>%
                 dplyr::select(Cell_Type, Marker) %>%
                 dplyr::rename(features.plot = Marker),
               text_size = 6, label = FALSE) + RotatedAxis() +
  ggtitle("All Cell Type Markers")

# plot side metrics
plot_dot_side(p, gex_multi, row_identity = "seurat_clusters",
              facet_col = "Cell_Type", info_to_add = c("cluster_size"))
```

```{r cluster-annot}
# annotations
annotations_general <- rbind(
  c("0", "CD4 Central Memory"),
  c("1", "CD4 Central Memory"),
  c("2", "CD14 Monocytes"),
  c("3", "CD4 Naive"),
  c("4", "NK cells"),
  c("5", "CD8 EM"),
  c("6", "CD8 EM"),
  c("7", "B cells"),
  c("8", "Low_quality_RNA_contamination_1"),  # low-quality/RNA contamination
  c("9", "Multiplets"),
  c("10", "CD8 EM"),
  c("11", "Treg"),
  c("12", "CD16 Monocytes"),
  c("13", "MAIT"),
  c("14", "CD8 Naive"),
  c("15", "cDC2"),
  c("16", "Low_quality_RNA_contamination_2"),      # low-quality/RNA contamination 
  c("17", "Low_quality_RNA_contamination_3"),      # low-quality/RNA contamination 
  c("18", "Microglia-like"),
  c("19", "Eosinophils/Basophils"),
  c("20", "pDC"),
  c("21", "Platelets"),
  c("22", "Plasmablasts"),
  c("23", "T_microglia"),     # possible doublets / RNA contamination
  c("24", "Proliferating"), 
  c("25", "cDC1"),
  c("26", "HPC"),
  c("27", "Erythrocytes")
)         
colnames(annotations_general) <- c("Cluster", "CellType")
annotations_general <- data.frame(annotations_general)

# create object to use in RenameIdents
annotated_clusters_all <- setNames(annotations_general$CellType, annotations_general$Cluster)

# Rename Idents
Idents(gex_multi) <- gex_multi$seurat_clusters
gex_multi <- RenameIdents(gex_multi, annotated_clusters_all)
gex_multi[["annotated_clusters_all"]] <- Idents(gex_multi)
gex_multi@meta.data$annotated_clusters_all <- factor(gex_multi@meta.data$annotated_clusters_all, levels = sort(as.character(unique(gex_multi@meta.data$annotated_clusters_all))))

# remove unnecessary objects
rm(annotations_general, annotated_clusters_all)
```

### Check for low quality cells and misclustering

```{r}
VlnPlot(gex_multi, 
        features = c("percent.mt", "percent.ribo", "nCount_RNA", "nFeature_RNA"), 
        group.by = "annotated_clusters_all", ncol = 1, pt.size = 0) + plot_layout(axes = "collect_x", axis_titles = "collect_x")
## some clusters are of low quality and therefore will be removed from further analysis
```

```{r}
# check for post-annotation batch effects
plot_pcts(pcts = calc_pcts(data = gex_multi[[]],
                           meta_group_by = c("annotated_clusters_all", "tissue"),
                           focus_group = "subject_id"), 
          tissue_type = "", plot_type = "All", plot_value = "Subject_id",
          count_by = c("annotated_clusters_all", "tissue"),
          x_axis = "annotated_clusters_all", x_axis_label = "Cell cluster", 
          fill_type = "subject_id", fill_label = "Subject ID", 
          details = "Resolution 0.8") +
  facet_wrap(~ tissue, ncol = 1) +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(size = 12)) + 
  labels_rotate_x
```

```{r}
# check for overlap with Azimuth annotations
table(gex_multi$annotated_clusters_all, gex_multi$predicted.celltype.l2)
```

### Filter final clusters and refine annotations

```{r}
# find subcluster
gex_multi_test <- FindSubCluster(gex_multi, 
                                 cluster = "T_microglia", subcluster.name = "annotated_clusters_all_ref", 
                                 resolution = 0.4, graph.name = "RNA_snn")

p <- DotPlot(gex_multi_test, cluster.idents = TRUE, group.by = "annotated_clusters_all_ref",
        features = unique(markers_all$Marker)) %>%
  add_info_bar(method = "join", 
               info_type = "Cell_Type",
               info = markers_all %>%
                 dplyr::select(Cell_Type, Marker) %>%
                 dplyr::rename(features.plot = Marker),
               text_size = 6, label = FALSE) + RotatedAxis() +
  ggtitle("All Cell Type Markers")

plot_dot_side(p, gex_multi_test, row_identity = "annotated_clusters_all_ref",
              facet_col = "Cell_Type", info_to_add = c("cluster_size"))
```

```{r}
# remove barcodes
gex_multi <- subset(gex_multi, subset = annotated_clusters_all %in% c("Multiplets", "T_microglia", "Low_quality_RNA_contamination_1",
                                                                      "Low_quality_RNA_contamination_2", "Low_quality_RNA_contamination_3"), invert = TRUE)

# rerun UMAP for better visualization after filtering
gex_multi <- RunUMAP(gex_multi, reduction = "harmony", dims = 1:50)

# add high-level annotation layer for basic cell types
gex_multi@meta.data <- gex_multi@meta.data %>%
                          mutate(annotated_clusters_gen = case_when(
                                          annotated_clusters_all %in% c("CD4 Central Memory", "CD4 Naive", "CD8 Naive", "Treg", "MAIT", "CD8 EM") ~ "T",
                                          annotated_clusters_all %in% c("CD14 Monocytes", "CD16 Monocytes") ~ "Mono",
                                          annotated_clusters_all %in% c("B cells", "Plasmablasts") ~ "B",
                                          annotated_clusters_all %in% c("cDC1", "cDC2", "pDC") ~ "DC",
                                          annotated_clusters_all %in% c("NK cells", "NK_CD56bright") ~ "NK",
                                          TRUE ~ annotated_clusters_all))
```

### Post-annotation quality check

```{r}
# plot UMAP with annotations
DimPlot(gex_multi, reduction = "umap.harmony", group.by = "annotated_clusters_all", 
              raster = FALSE, label = TRUE, label.box = TRUE, repel = TRUE) + 
  scale_fill_manual(values = rep("#FFFFFF", length(gex_multi$annotated_clusters_all))) +
  NoLegend()

# check for differential cell types based on tissue and HIV infection
p1 <- DimPlot(gex_multi, reduction = "umap.harmony", split.by = "tissue", raster = FALSE, label = FALSE) 
p2 <- DimPlot(gex_multi, reduction = "umap.harmony", split.by = "HIV_status", raster = FALSE, label = FALSE, repel = TRUE)
p1 / p2 + plot_layout(guides = "collect")

# cluster barplots for quality check
plot_pcts(pcts = calc_pcts(data = gex_multi[[]],
                           meta_group_by = c("annotated_clusters_all", "tissue"),
                           focus_group = "subject_id"), 
          tissue_type = "", plot_type = "All", plot_value = "Subject_id",
          count_by = c("annotated_clusters_all", "tissue"),
          x_axis = "annotated_clusters_all", x_axis_label = "Cell cluster", 
          fill_type = "subject_id", fill_label = "Subject ID", 
          details = "Resolution 0.8") +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(size = 12)) + 
  facet_grid(~ tissue, scales = "free") +
  labels_rotate_x

plot_pcts(pcts = calc_pcts(data = gex_multi[[]],
                           meta_group_by = c("annotated_clusters_all", "tissue"),
                           focus_group = "HIV_status"), 
          tissue_type = "", plot_type = "All", plot_value = "HIV Status",
          count_by = c("annotated_clusters_all", "tissue"),
          x_axis = "annotated_clusters_all", x_axis_label = "Cell cluster", 
          fill_type = "HIV_status", fill_label = "HIV Status", 
          details = "Resolution 0.5") +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(size = 12)) + 
  facet_wrap(~ tissue, nrow = 2, scales = "free") +
  labels_rotate_x

# Dot plot
p <- DotPlot(gex_multi, cluster.idents = TRUE, group.by = "annotated_clusters_all",
        features = unique(markers_all$Marker)) %>%
  add_info_bar(method = "join", 
               info_type = "Cell_Type",
               info = markers_all %>%
                 dplyr::select(Cell_Type, Marker) %>%
                 dplyr::rename(features.plot = Marker),
               text_size = 6, label = FALSE) + RotatedAxis() +
  ggtitle("All Cell Type Markers")

plot_dot_side(p, gex_multi, row_identity = "annotated_clusters_all",
              facet_col = "Cell_Type", info_to_add = c("cluster_size"))
```
